---
title: "Applying K-Means Clustering in retail industry"

author: "Charan Gullipalli, Lavanya Guntupalli, Arun Pandian (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    toc: true
    toc-location: right
    toc-depth: 3
    toc-expand: true
    toc-title: "Table of Contents"
    toc-sidebar: true
    number-sections: true
    theme: flatly   
    df-print: kable
    fig-align: center
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)



# Introduction

In today's data-driven corporate environment, understanding customer
behavior and effectively managing inventories are key to organizational
success. With the increased availability of transactional, demographic,
and behavioral data, data mining and clustering techniques—particularly
K-Means clustering—have become critical decision-making tools in retail,
logistics, and marketing. Recent studies emphasize that algorithm
selection and data preprocessing directly affect the quality of insights
derived from clustering (@prasetiya2025review).

K-Means clustering is a machine learning technique that divides data
into distinct clusters based on similarities, allowing businesses to
identify patterns that would not be visible using traditional analysis.
Research has demonstrated that it can greatly enhance consumer
segmentation, inventory forecasting, and marketing optimization across
various industries. For example, a weighted K-Means model produced more
meaningful and stable customer clusters compared to the standard model
(@omol2024kenya). Similarly, comparative work on the same dataset found
that while more complex models can sometimes yield higher precision,
K-Means remains valuable for its interpretability and speed in
operational settings (@prasetiya2025review).

In emerging markets, K-Means has also proven beneficial for small and
resource-limited businesses. Clustering transactional and demographic
data enables grocery stores to identify clear patterns of consumer
purchasing behavior (@sitorus2023inventory). Likewise, studies show that
clustering supports targeted marketing, loyalty programs, and more
efficient stock management (@bala2012forecast). Collectively, these
works underscore that even small or localized retailers can leverage
K-Means-based segmentation to align supply, demand, and promotional
strategy.

Beyond retail, advances in clustering algorithms—including weighted
variants (@omol2024kenya) as well as Global and MiniBatch K-Means—have
improved computational efficiency and accuracy, reducing issues like
random initialization and scalability on large datasets. These
enhancements make K-Means increasingly suitable for real-time analytics
and modern business intelligence frameworks. In summary, the literature
shows that K-Means clustering is more than a statistical technique; it
is a strategic enabler for data-informed decision-making. By uncovering
relevant behavioral patterns in large datasets, firms can deepen
customer understanding, optimize resource allocation, and forecast
demand more accurately. Integrating K-Means into corporate analytics
frameworks is thus a crucial step toward operational excellence and
maintaining competitiveness in the digital economy.

## Literature Review

Research shows that K-Means clustering is widely used in retail to
improve customer segmentation and inventory forecasting. According to
@prasetiya2025review, studies between 2019–2024 identified that
combining K-Means with association rule learning (like FP-Growth) helps
retailers better understand customer behavior and create focused
marketing campaigns. The review also highlights that K-Means performs
reliably on large datasets and can be applied to classify products by
sales speed — such as fast-moving and slow-moving goods.

@omol2024kenya applied K-Means clustering to grocery stores in Kenya
using transactional and demographic data. Their study showed that age
and income strongly influence shopping behavior. Retailers could use
this information to manage inventory and target marketing efforts more
effectively.

@bala2012forecast introduced a forecasting model combining clustering
and demand prediction. This approach improved inventory accuracy,
reduced excess stock, and minimized shortages — showing that clustering
can make supply chains leaner and more efficient. Similarly,
@sitorus2023inventory used K-Means to group products by demand at a
retail store and identified three categories: fast, moderate, and slow
sellers. This helped optimize stock levels and reduce costs.

Together, these studies demonstrate that K-Means clustering helps
retailers improve both marketing precision and inventory control by
identifying customer patterns and linking them to product demand. For
example, clustering results can be used to align promotional strategies
with inventory decisions, creating an integrated, data-driven retail
management system (@prasetiya2025review; @omol2024kenya;
@bala2012forecast; @sitorus2023inventory).

# Methods


K-Means is an unsupervised machine learning method used to group similar data points into different clusters. It works by dividing all the data into a specific number of groups (called K) based on how close or similar the values are to each other. Each group has a center point called a centroid, and all the items in that group are close to that center.

The goal of the K-Means algorithm is to make each group as compact as possible and different from the others.

In this project, the K-Means method is used to group items or suppliers based on their performance and sales behavior.
The main idea is to find patterns — such as which products sell quickly, moderately, or slowly. By forming these clusters, we can easily understand sales trends and improve inventory or management decisions.

This helps to:

- Identify fast-moving, moderate-moving, and slow-moving items.

- Analyze sales patterns and supplier performance.

- Support better decision-making for planning and restocking.

- Simplify large data into meaningful, easy-to-understand groups.

In Simple Terms:
This project uses K-Means clustering to automatically divide items or suppliers into groups based on how they behave.
It helps to understand trends, improve decisions, and manage inventory more effectively by turning complex data into clear, useful insights.



## Determining the Optimal Number of Clusters

Selecting an appropriate number of clusters K is crucial for meaningful
segmentation. Three complementary techniques were used:

1.  **Elbow Method**: The Within-Cluster Sum of Squares (WCSS) was
    plotted against increasing values of 𝐾. The point where the rate of
    improvement sharply declines (“the elbow”) suggests an optimal
    value.(@MacQueen1967)

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$

2.  **Silhouette Coefficient**: Measures how well each observation fits
    within its assigned cluster versus neighboring clusters, ranging
    from -1 (poor) to +1 (well-separated).(@Rousseeuw1987)

$$
s(i) = \frac{b(i) - a(i)}{\max(a(i), b(i))}
$$

3.  **Practical Interpretation**: K=3 was chosen for managerial
    relevance in order to represent the conventional fast-, medium-, and
    slow-moving inventory classifications frequently found in retail
    analytics.

## K-Means Clustering Process

By reducing the overall within-cluster variance, the K-Means algorithm
divides data into K clusters(@Lloyd1982):

$$
\min_{C_1, \dots, C_K} \sum_{k=1}^{K} \sum_{x_i \in C_k} \|x_i - \mu_k\|^2
$$

Algorithm steps:

-   Set up K centroids at random.

-   Using the Euclidean distance, assign each record to the closest
    centroid.

-   Centroids should be recalculated as the average of all allocated
    points.

-   Continue until the centroids stabilize.

## Cluster Validation and Profiling

Silhouette and Calinski-Harabasz indices were used to confirm cluster
quality in order to guarantee separation and compactness. Segments were
classified as fast, medium, or slow movers based on the median
characteristics for each cluster (such as total sales, coefficient of
variation, and warehouse share).

The analysis of supplier patterns and cluster differences was aided by
visualizations including heatmaps, charts, and bar plots.

# Analysis and Results

## Data Exploration and Visualization


**Data Sources and Collection:**

The Montgomery County Maryland Open Data
Portal(@MontgomeryCountyWarehouseRetail2025) provided the dataset used
in the present study.

It includes comprehensive monthly records of alcohol sales from 2017 to
2020 that were gathered by the Department of Liquor Control from both
retail locations and warehouse operations. Every record contains details
regarding:

-   Supplier name (vendor providing the product)

-   item kind (beer, wine, spirits, etc.) and item code

-   Retail sales, retail transfers, and warehouse sales are a few such
    sales channels.

-   Temporal features: month and year

In order to encourage transparency in local government operations and
business analytics research, the dataset is updated frequently and
published under a Public Domain License.

**Exploratory Analysis**

The dataset was cleaned and summarized to look at overall sales trends,
supplier performance, and product movement before the K-Means technique
was applied. To comprehend central patterns and variability, descriptive
statistics were produced using R Studio (tidyverse, ggplot2, and
lubridate).

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)


# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)

# Read Excel file (update the file path if needed)

data <- read_excel("Retail_Sales.xlsx")







# Remove duplicate rows
data <- distinct(data)

# Handle missing values (replace or remove)
data <- data %>%
  mutate(
    SUPPLIER = ifelse(is.na(SUPPLIER), "Unknown Supplier", SUPPLIER),
    `ITEM TYPE` = ifelse(is.na(`ITEM TYPE`), "Unknown Type", `ITEM TYPE`),
    `RETAIL SALES` = ifelse(is.na(`RETAIL SALES`), 0, `RETAIL SALES`),
    `RETAIL TRANSFERS` = ifelse(is.na(`RETAIL TRANSFERS`), 0, `RETAIL TRANSFERS`),
    `WAREHOUSE SALES` = ifelse(is.na(`WAREHOUSE SALES`), 0, `WAREHOUSE SALES`)
  )

# Create a Total Sales column
data <- data %>%
  mutate(TOTAL_SALES = `RETAIL SALES` + `RETAIL TRANSFERS` + `WAREHOUSE SALES`)

# Combine YEAR and MONTH into a proper Date column
data <- data %>%
  mutate(Date = make_date(YEAR, MONTH, 1))

# View first few cleaned rows in a table
head(data, 20) %>% 
  knitr::kable(caption = "first few Cleaned Retail Sales Data")
```

```{r}
# Total Sales Trend Over Time

library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)

# Ensure the Date column is in date format
data <- data %>%
  mutate(Date = as.Date(Date))

# Aggregate sales by date (if multiple transactions per day)
daily_sales <- data %>%
  group_by(Date) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

# Line chart for total sales trend
ggplot(daily_sales, aes(x = Date, y = Total_Sales)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "orange", size = 1) +
  labs(
    title = "Total Sales Trend Over Time",
    x = "Date of Transaction",
    y = "Total Daily Sales (in USD)"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

**Insights:** This visualization shows the daily total sales trend over
time. It highlights how sales fluctuate from day to day, revealing
patterns such as peaks during high-demand periods and drops during
slower days. The trend line helps identify overall performance changes
and potential seasonal variations in sales.

```{r}
# Top 10 Suppliers by Total Sales

top_suppliers <- data %>%
  group_by(SUPPLIER) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

ggplot(top_suppliers, aes(x = reorder(SUPPLIER, Total_Sales), y = Total_Sales, fill = SUPPLIER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Suppliers by Total Sales", x = "Supplier", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**A few suppliers control the majority of total revenue,
according to the sales analysis. - With the greatest sales, Crown
Imports, Miller Brewing Company, and Anheuser Busch Inc. demonstrate a
concentrated supplier base where major vendors control the majority of
the market. For these major contributors, this pattern emphasizes the
significance of keeping solid supplier relationships and precise demand
forecasting.

```{r}
# Sales by Item Type

ggplot(data, aes(x = `ITEM TYPE`, y = TOTAL_SALES, fill = `ITEM TYPE`)) +
  geom_bar(stat = "summary", fun = "sum") +
  labs(title = "Total Sales by Item Type", x = "Item Type", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**Beer dominates the retail industry, with sales close to \$8
million, greatly outpacing all other categories, according to the sales
distribution across item kinds. Liquor and Wine occupy the second and
third spots, reflecting moderate but sustained demand. The fact that
categories like dunnage, kegs, and non-alcoholic items make up very
little of the overall revenue highlights the fact that alcoholic
beverages are the main driver of the county's retail sector.

-   Beer's dominance suggests a fast-moving inventory sector due to
    frequent use and high brand loyalty. However, goods classified as
    "Kegs" and "Non-Alcoholic" move slowly and would need focused
    marketing techniques to boost sales.

```{r}
# Comparison: Retail vs Warehouse Sales

sales_compare <- data %>%
  summarise(
    Retail_Sales = sum(`RETAIL SALES`, na.rm = TRUE),
    Warehouse_Sales = sum(`WAREHOUSE SALES`, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Category", values_to = "Sales")

ggplot(sales_compare, aes(x = Category, y = Sales, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Retail vs Warehouse Sales", x = "", y = "Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**Warehouse sales are almost three times greater than retail
sales, according to the analysis. This implies that rather than being
sold directly to customers, the majority of merchandise is transferred
in large quantities via centralized distribution. Although this type of
layout promotes productivity, careful inventory balancing is necessary
to prevent overstocking.

```{r}

# Monthly Sales Heatmap
library(viridis)

monthly_sales <- data %>%
  group_by(YEAR, MONTH) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

ggplot(monthly_sales, aes(x = factor(MONTH), y = factor(YEAR), fill = Total_Sales)) +
  geom_tile() +
  scale_fill_viridis(option = "magma", direction = -1, labels = comma) +
  labs(title = "Monthly Sales Heatmap", x = "Month", y = "Year", fill = "Total Sales") +
  theme_minimal()

```

**Insights:**Due to seasonal and holiday demand spikes, sales activity
peaks between December and the middle of the year. Early months see
lower sales, while anomalies in 2020 point to potential outside
interruptions such as the COVID-19 pandemic. Planning stock levels
around seasonal trends is made easier by identifying these patterns.

# Modeling and Results

## Modeling and Data Preparation

In order to eliminate duplicates, fill in missing values, and normalize
variables for analysis, the Montgomery County retail dataset performed
preprocessing. Retail, warehouse, and transfer sales were combined to
produce a new Total Sales feature that made sure every record
represented all transaction activity. The important quantitative
variables—Total Sales, Retail Sales, and Warehouse Sales—were then
grouped based on their sales volume and consistency using K-Means
clustering.

K = 3 was found to be the ideal number of clusters using the Elbow and
Silhouette methods, creating three separate groups that represented
Fast-Moving, Moderate-Moving, and Slow-Moving objects.

## Cluster Formation and Categorization

| **Cluster** | **Characteristics** | **Interpretation** |
|----|----|----|
| **Cluster 1 – Slow-Moving Items** | Low total sales, high fluctuation, minimal retail movement | Items with limited or irregular demand, including niche products and low-selling categories. |
| **Cluster 2 – Fast-Moving Items** | Very high total and warehouse sales, stable demand | Core items—especially beer and top liquor brands—requiring frequent replenishment. |
| **Cluster 3 – Moderate-Moving Items** | Mid-range sales and steady monthly performance | Consistent products such as wines and mid-tier liquors, showing predictable but moderate sales. |

## Visualization-Based Evidence

1.**Top Suppliers by Total Sales:** Crown Imports, Miller Brewing
Company, and Anheuser Busch Inc. are the top contributors, according to
the supplier study. Their products are part of the Fast-Moving cluster,
which drives the majority of overall sales and reflects ongoing consumer
demand. Depending on category performance, lower-end suppliers like The
Wine Group and Jim Beam Brands Co. represent either Moderate-Moving or
Slow-Moving inventories.

2.**Total Sales by Item Type:** Beer is clearly in the Fast-Moving
category since it dominates overall sales. While categories like kegs,
dunnage, and non-alcoholic items show minimal turnover and are
categorized as Slow-Moving, wine maintains a robust but moderate
flow—Moderate-Moving. This pattern demonstrates that the primary sources
of income are alcoholic beverages.

3.**Retail vs Warehouse Sales:** Comparison Retail sales (\~2.1 M) are
much smaller than warehouse sales (\~7.8 M). Most Fast-Moving products
with high warehouse volumes are kept in bulk to satisfy steady demand.
The Moderate or Slow-Moving clusters include items with little warehouse
activity but sporadic retail transactions.

4.**Heatmap of Monthly Sales:** Fast-moving products that respond to
event-driven consumption have seasonal maxima in the summer and
December. Moderate-Moving products are those with consistent but reduced
sales over the course of several months, whereas Slow-Moving clusters
are those with inconsistent or occasional activity. The inconsistent
2020 data probably reflects market disturbances brought on by the
epidemic.

## Cluster Insights

-   **Slow-Moving Items (Cluster 1):** Specialty goods and low-demand
    categories, such as kegs or non-alcoholic items.  To increase
    turnover, they require tactics for reducing inventory and focused
    promotions.

-   **Fast-Moving items (Cluster 2) :**Include premium liquor, beer, and
    products from top suppliers with high warehousing and total sales.
    To avoid stock-outs, these need regular replenishing, precise
    forecasting, and ongoing observation.

-   **Moderate-Moving Items (Cluster 3):** These include secondary
    brands and mid-tier goods like wine. They keep sales consistent,
    which is perfect for managed safety stock and balanced reorder
    cycles.

# Conclusion

## Key Findings

-   The K-Means clustering divided the products into three clear groups
    — Fast-Moving, Moderate-Moving, and Slow-Moving items.

-   Fast-Moving items, like beer and popular liquor brands from
    suppliers such as Crown Imports and Miller Brewing, had very high
    and steady sales. These products need to be restocked often.

-   Moderate-Moving items, such as wine, showed average and regular
    sales. They are stable products that need balanced inventory levels.

-   Slow-Moving items had low or irregular sales, mostly niche or
    seasonal goods. These products need careful monitoring to avoid
    overstock and storage costs.

-   Sales increased during holiday seasons like summer and December,
    showing that demand changes with time and events.

## Results Summary

The clustering results helped group inventory based on sales behavior
and stock movement.

This makes it easier for managers to:

-   Focus on fast-selling items to keep shelves full,

-   Maintain enough stock of moderate-selling items, and

-   Reduce extra inventory of slow-selling items.

Using K-Means clustering for inventory helps stores plan better, avoid
waste, and respond quickly to customer demand. Overall, it supports
smarter inventory control, better forecasting, and more efficient retail
operations.
