---
title: "Applying K-Means Clustering in retail industry - Fall 2025 "

author: "Charan Gullipalli, Lavanya Guntupalli, Arun Pandian (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction

## 1. Introduction

In today's data-driven corporate environment, understanding customer
behavior and effectively managing inventories are key to organizational
success. With the increased availability of transactional, demographic,
and behavioral data, data mining and clustering techniques—particularly
K-Means clustering—have become critical decision-making tools in retail,
logistics, and marketing. Recent studies emphasize that algorithm
selection and data preprocessing directly affect the quality of insights
derived from clustering (@prasetiya2025review).

K-Means clustering is a machine learning technique that divides data
into distinct clusters based on similarities, allowing businesses to
identify patterns that would not be visible using traditional analysis.
Research has demonstrated that it can greatly enhance consumer
segmentation, inventory forecasting, and marketing optimization across
various industries. For example, a weighted K-Means model produced more
meaningful and stable customer clusters compared to the standard model
(@omol2024kenya). Similarly, comparative work on the same dataset found
that while more complex models can sometimes yield higher precision,
K-Means remains valuable for its interpretability and speed in
operational settings (@prasetiya2025review).

In emerging markets, K-Means has also proven beneficial for small and
resource-limited businesses. Clustering transactional and demographic
data enables grocery stores to identify clear patterns of consumer
purchasing behavior (@sitorus2023inventory). Likewise, studies show that
clustering supports targeted marketing, loyalty programs, and more
efficient stock management (@bala2012forecast). Collectively, these
works underscore that even small or localized retailers can leverage
K-Means-based segmentation to align supply, demand, and promotional
strategy.

Beyond retail, advances in clustering algorithms—including weighted
variants (@omol2024kenya) as well as Global and MiniBatch K-Means—have
improved computational efficiency and accuracy, reducing issues like
random initialization and scalability on large datasets. These
enhancements make K-Means increasingly suitable for real-time analytics
and modern business intelligence frameworks. In summary, the literature
shows that K-Means clustering is more than a statistical technique; it
is a strategic enabler for data-informed decision-making. By uncovering
relevant behavioral patterns in large datasets, firms can deepen
customer understanding, optimize resource allocation, and forecast
demand more accurately. Integrating K-Means into corporate analytics
frameworks is thus a crucial step toward operational excellence and
maintaining competitiveness in the digital economy.

## 2. Literature Review

Research shows that K-Means clustering is widely used in retail to
improve customer segmentation and inventory forecasting. According to
@prasetiya2025review, studies between 2019–2024 identified that
combining K-Means with association rule learning (like FP-Growth) helps
retailers better understand customer behavior and create focused
marketing campaigns. The review also highlights that K-Means performs
reliably on large datasets and can be applied to classify products by
sales speed — such as fast-moving and slow-moving goods.

@omol2024kenya applied K-Means clustering to grocery stores in Kenya
using transactional and demographic data. Their study showed that age
and income strongly influence shopping behavior. Retailers could use
this information to manage inventory and target marketing efforts more
effectively.

@bala2012forecast introduced a forecasting model combining clustering
and demand prediction. This approach improved inventory accuracy,
reduced excess stock, and minimized shortages — showing that clustering
can make supply chains leaner and more efficient. Similarly,
@sitorus2023inventory used K-Means to group products by demand at a
retail store and identified three categories: fast, moderate, and slow
sellers. This helped optimize stock levels and reduce costs.

Together, these studies demonstrate that K-Means clustering helps
retailers improve both marketing precision and inventory control by
identifying customer patterns and linking them to product demand. For
example, clustering results can be used to align promotional strategies
with inventory decisions, creating an integrated, data-driven retail
management system (@prasetiya2025review; @omol2024kenya;
@bala2012forecast; @sitorus2023inventory).

## Methods

In this project, customers are grouped according to their purchase
habits in the retail dataset using the K-Means clustering technique.
Finding market niches like big spenders, infrequent consumers, and
frequent buyers is the goal.

Prior to modeling, duplicates were eliminated, missing values were
handled, and new fields like total amount (Quantity × Unit Price) were
created. After then, the data was standardized so that every variable
could be compared.

Customer behavior was represented using the RFM model (Recency,
Frequency, Monetary). The K-Means algorithm was fed these three
characteristics, and it operates by:

-   selecting K cluster centers at random.

-   Euclidean distance is used to assign each data point to the closest
    center.

-   utilizing the mean of the allotted points to update the centers.

-   Continue until the centers stop shifting.

The optimal number of clusters (K) was determined using the Elbow method
and Silhouette score.

Because K-Means is effective, simple to understand, and appropriate for
large-scale retail data, it was chosen. The findings aid in
comprehending various client categories for focused marketing tactics.

## Analysis and Results

# Data Exploration and Visualization

R Studio was used to clean and visualize the data. Monthly trends,
supplier performance, and product categories were the main subjects of
the analysis.

The holidays are when sales are at their highest, according to the
visuals, and a select group of suppliers generate the greatest income.
The best-selling products were wine and spirits, and warehouse sales
were typically higher than retail sales.

These trends demonstrate the dataset's substantial provider dominance
and seasonal demand.

Data Sources and Collection:

-   The Department of Liquor Control's Montgomery County, Maryland Open
    Data Portal provided the data. Monthly retail, transfer, and
    warehouse sales records by supplier and item type are included;
    these records are updated on a regular basis and made publicly
    accessible under a Public Domain License.

Initial Findings and Insights:

-   Holiday seasons are when sales are at their highest.

-   Overall sales are dominated by a small number of vendors.

-   Wine and spirits are the most popular product categories.

-   Sales in warehouses are higher than those in stores.

Unexpected Patterns or Anomalies:

-   There are also duplicate item codes and missing supplier names.

-   certain months with sales figures that are abnormally high or low.

-   Retail transfer increases are probably related to restocking
    occasions.

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)


# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)

# Read Excel file (update the file path if needed)

data <- read_excel("Retail_Sales.xlsx")

# View basic structure
glimpse(data)
summary(data)




# Remove duplicate rows
data <- distinct(data)

# Handle missing values (replace or remove)
data <- data %>%
  mutate(
    SUPPLIER = ifelse(is.na(SUPPLIER), "Unknown Supplier", SUPPLIER),
    `ITEM TYPE` = ifelse(is.na(`ITEM TYPE`), "Unknown Type", `ITEM TYPE`),
    `RETAIL SALES` = ifelse(is.na(`RETAIL SALES`), 0, `RETAIL SALES`),
    `RETAIL TRANSFERS` = ifelse(is.na(`RETAIL TRANSFERS`), 0, `RETAIL TRANSFERS`),
    `WAREHOUSE SALES` = ifelse(is.na(`WAREHOUSE SALES`), 0, `WAREHOUSE SALES`)
  )

# Create a Total Sales column
data <- data %>%
  mutate(TOTAL_SALES = `RETAIL SALES` + `RETAIL TRANSFERS` + `WAREHOUSE SALES`)

# Combine YEAR and MONTH into a proper Date column
data <- data %>%
  mutate(Date = make_date(YEAR, MONTH, 1))

```

```{r}
# Total Sales Trend Over Time

ggplot(data, aes(x = Date, y = TOTAL_SALES)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(title = "Total Sales Trend Over Time", x = "Date", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

```{r}
# Top 10 Suppliers by Total Sales

top_suppliers <- data %>%
  group_by(SUPPLIER) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

ggplot(top_suppliers, aes(x = reorder(SUPPLIER, Total_Sales), y = Total_Sales, fill = SUPPLIER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Suppliers by Total Sales", x = "Supplier", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
# Sales by Item Type

ggplot(data, aes(x = `ITEM TYPE`, y = TOTAL_SALES, fill = `ITEM TYPE`)) +
  geom_bar(stat = "summary", fun = "sum") +
  labs(title = "Total Sales by Item Type", x = "Item Type", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
# Comparison: Retail vs Warehouse Sales

sales_compare <- data %>%
  summarise(
    Retail_Sales = sum(`RETAIL SALES`, na.rm = TRUE),
    Warehouse_Sales = sum(`WAREHOUSE SALES`, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Category", values_to = "Sales")

ggplot(sales_compare, aes(x = Category, y = Sales, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Retail vs Warehouse Sales", x = "", y = "Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}

# Monthly Sales Heatmap
library(viridis)

monthly_sales <- data %>%
  group_by(YEAR, MONTH) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

ggplot(monthly_sales, aes(x = factor(MONTH), y = factor(YEAR), fill = Total_Sales)) +
  geom_tile() +
  scale_fill_viridis(option = "magma", direction = -1, labels = comma) +
  labs(title = "Monthly Sales Heatmap", x = "Month", y = "Year", fill = "Total Sales") +
  theme_minimal()

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.
