---
title: "Writing a great story for data science projects - Fall 2025 "
subtitle: "This is a Report Template Quarto"
author: "Charan (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
:::

## Introduction


In today's data-driven corporate environment, understanding customer behavior and effectively managing inventories are key to organizational success. With the increased availability of transactional, demographic, and behavioral data, data mining and clustering techniques—particularly K-Means clustering—have become critical decision-making tools in retail, logistics, and marketing. Recent studies emphasize that algorithm selection and data preprocessing directly affect the quality of insights derived from clustering (Prasetiya and Prayoga 2024).
K-Means clustering is a machine learning technique that divides data into distinct clusters based on similarities, allowing businesses to identify patterns that would not be visible using traditional analysis. Research has demonstrated that it can greatly enhance consumer segmentation, inventory forecasting, and marketing optimization across various industries. For example, a weighted K-Means model produced more meaningful and stable customer clusters compared to the standard model (Omol et al. 2024). Similarly, comparative work on the same dataset found that while more complex models can sometimes yield higher precision, K-Means remains valuable for its interpretability and speed in operational settings (Prasetiya and Prayoga 2024).
In emerging markets, K-Means has also proven beneficial for small and resource-limited businesses. Clustering transactional and demographic data enables grocery stores to identify clear patterns of consumer purchasing behavior (Sitorus et al. 2023). Likewise, studies show that clustering supports targeted marketing, loyalty programs, and more efficient stock management (Bala 2012; Sitorus et al. 2023; Gunawan 2022). Collectively, these works underscore that even small or localized retailers can leverage K-Means-based segmentation to align supply, demand, and promotional strategy.
Beyond retail, advances in clustering algorithms—including weighted variants (Omol et al. 2024) as well as Global and MiniBatch K-Means—have improved computational efficiency and accuracy, reducing issues like random initialization and scalability on large datasets. These enhancements make K-Means increasingly suitable for real-time analytics and modern business intelligence frameworks.
In summary, the literature shows that K-Means clustering is more than a statistical technique; it is a strategic enabler for data-informed decision-making. By uncovering relevant behavioral patterns in large datasets, firms can deepen customer understanding, optimize resource allocation, and forecast demand more accurately. Integrating K-Means into corporate analytics frameworks is thus a crucial step toward operational excellence and maintaining competitiveness in the digital economy.


The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

<!-- -->

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

Example of writing including citing references:

*This is an introduction to ..... regression, which is a non-parametric
estimator that estimates the conditional expectation of two variables
which is random. The goal of a kernel regression is to discover the
non-linear relationship between two random variables. To discover the
non-linear relationship, kernel estimator or kernel smoothing is the
main method to estimate the curve for non-parametric statistics. In
kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]*. Topology can be used in machine
learning [@adams2021topology]

For Symbolic Regression [@wang2019symbolic] *This is my work and I want
to add more work...*

Cite new paper [@su2012linear]

#Literature



- Customer segmentation through clustering techniques has become essential for retailers to understand consumer behavior and improve marketing and inventory decisions. K-Means clustering is still a popular technique among these due to its ease of use, interpretability, and effectiveness when working with big datasets. The following studies show how K-Means has been successfully used in a variety of retail settings, including department shops, supermarkets, and online platforms.


- K-Means was utilized in another study on a department store example to create user profiles using membership and transaction data. The authors identified ideal clusters that reflected important shopper kinds, such as high-value regulars and sporadic low spenders, using the elbow approach and the Calinski–Harabasz index. Although the study highlighted the importance of clustering for targeted marketing and loyalty programs, it lacked behavioral or demographic diversity, which prevented a more thorough understanding of consumer motives.

- A study that used a weighted K-Means version with RFM (Recency, Frequency, Monetary) analysis aimed to improve segmentation accuracy in the online retail industry. The weighted version outperformed the regular K-Means (around 0.30) in terms of silhouette scores using a sizable UK retail dataset (about 0.40). More useful client clusters for targeting and retention were produced by this method. However, it overlooked other behavioral elements like browsing or engagement patterns because it just used RFM data.

- Several clustering techniques, including K-Means, Gaussian Mixture Models (GMM), DBSCAN, BIRCH, and Agglomerative clustering, were investigated using RFM-based features in a comparative analysis on UK retail data. The results indicated that K-Means remained the most practical due to its simplicity and quick computation, but GMM performed the best (silhouette score ~0.80). The study came to the conclusion that segmentation quality is greatly impacted by algorithm choice, and that results might be further improved with richer datasets.

- All things considered, the examined literature suggests K-Means clustering is a dependable and flexible method for retail customer segmentation. It enables data-driven marketing, inventory, and customer relationship management initiatives by assisting organizations in meaningfully grouping their customers. Although the majority of studies still only use transactional data, those that incorporate weighted features or RFM analysis show increased accuracy. In order to obtain more thorough and useful segmentation findings, future studies should investigate hybrid models and incorporate behavioral and demographic data.






Arun Pandian Madasamyraju:
1. Implementation of K-Means Clustering for Inventory Projection
Link: https://ijstm.inarah.co.id/index.php/ijstm/article/view/856
This study used K-means clustering to improve inventory forecasting accuracy in a trading firm (Dapur Uniq), grouping items with similar demand trends. Using Orange Data Mining for analysis, the researchers discovered three product clusters—fast-, medium-, and slow-moving items—to assist the organization in balancing stock levels and minimizing shortages and overstock. K-means was successful for simplifying inventory choices, optimizing replenishment cycles, and increasing supply chain efficiency; nevertheless, more study with bigger datasets is recommended to increase model generalizability.
2. Enhancing Customer Segmentation: RFM Analysis and K-Means Clustering Implementation
Link:https://www.researchgate.net/publication/358285794_Analysis_of_RFM_Customer_Segmentation_Using_Clustering_Algorithms


This article investigated the use of Recency, Frequency, and Monetary (RFM) analysis in conjunction with K-means clustering to increase client segmentation and marketing customization. By preprocessing retail transaction data and using K-means to RFM ratings, the study classified customers as "Best Customers" or "Almost Lost." This connectivity made it possible to run targeted promotions and manage client relationships more efficiently. However, the study concluded that segmentation success is dependent on data quality and the accurate selection of the appropriate K value, and it suggested future research on dynamic, real-time segmentation algorithms for more flexibility.
3. The Global K-Means Algorithm (Likas, Vlassis, & Verbeek)
PDF https://www.cs.uoi.gr/~arly/full/conferences/C53.pdf cs.uoi.gr
Likas et al. proposed the Global K-Means method as an incremental and deterministic enhancement over ordinary K-means to address its sensitivity to random initialization. This approach creates clusters one at a time from carefully selected beginning locations, assuring stability and consistency independent of original centroid choices. In benchmark datasets such as Iris and synthetic picture data, the technique produced decreased clustering error and improved resilience. The research demonstrated how global K-means provides a more consistent and precise clustering basis for applications such as picture segmentation and customer data analysis, which need repeatability and stability.
4. Analysis of Simple K-Means Clustering Algorithm (Kanungo et al.)
Kanungo and colleagues investigated the computational characteristics of the classic K-means method and suggested an improved filtering strategy that use kd-trees to increase performance and scalability. Their data-sensitive technique lowers superfluous distance computations while increasing convergence efficiency, especially when clusters are well-separated. Experimental findings on simulated and real datasets, including color quantization and picture compression, demonstrated that the updated approach achieves comparable accuracy while computing quicker. This makes K-means more suitable for large-scale data mining, segmentation, and compression applications requiring high processing efficiency.




Charan Gullipalli
1. Application of K-Means Clustering for Customer Segmentation in Grocery Stores in Kenya
 https://ijstm.inarah.co.id/index.php/ijstm/article/view/1024
Omol, Onyangor, Mburu, and Abuonji (2024) used K-means clustering on Kenyan grocery store data to identify customer groups with similar purchase tendencies based on demographic and transactional factors. The findings suggested that clustering successfully identified relevant customer categories, which guided focused marketing, improved service design, and increased customer happiness. K-means assists retailers in transforming raw purchase data into actionable customer insights, allowing grocery stores to focus on high-value categories and tailor offers to boost loyalty and sales.

2. Digital Maturity Action Fields for SMEs in Developing Economies
 https://www.researchgate.net/publication/372365772_DIGITAL_MATURITY_ACTION_FIELDS_FOR_SMEs_IN_DEVELOPING_ECONOMIES
Although not directly related to clustering, Omol, Mburu, and Abuonji's (2023) research explores how using digital technologies increases productivity and competitiveness in small and medium-sized firms (SMEs). K-means clustering can supplement these findings by evaluating digital adoption rates and categorizing SMEs based on technology preparedness, innovative capability, and workforce skill gaps. Such segmentation enables policymakers and managers to prioritize digital transformation plans and efficiently distribute resources across company clusters.
3. Recency, Frequency, Monetary Value, Clustering, and Internal and External Indices for Customer Segmentation from Retail Data
 https://www.mdpi.com/1999-4893/16/9/396
Wilbert (2023) used Recency-Frequency-Monetary (RFM) analysis together with K-means, K-medoids, and MiniBatch K-means clustering to discover six separate client categories, including loyal, lost, and best customers. K-means gave the most stable segmentation and assisted merchants in understanding purchase recency and spending value patterns, hence enabling tailored marketing and retention strategies. This demonstrates how K-means simplifies large-scale RFM data while identifying high-value clients for targeted interaction.

4. Improving Inventory Performance with Clustering-Based Demand Forecasts
 https://www.researchgate.net/publication/261023207_Decision_tree_based_demand_forecasts_for_improving_inventory_performance
Bala (2012) showed that clustering-based forecasting improves supply chain efficiency by grouping comparable demand trends across SKUs. The study found that using K-means in inventory models enhanced demand forecasting accuracy, lowered holding costs, and raised service levels when compared to traditional techniques. K-means is therefore useful for retailers and manufacturers looking to incorporate consumer segmentation into predictive models for more efficient inventory control and replenishment choices.
5. Comparative Analysis of Store Clustering Techniques in the Retail Industry
 https://www.scitepress.org/Papers/2019/79175/79175.pdf
Agarwal, Jain, and Rajnayak (2019) compared clustering methods—including hierarchical, density-based, and K-means—to analyze store-level performance and customer behavior. Their findings suggested that K-means efficiently unites retail outlets with comparable sales trends, allowing for more informed pricing, layout, and merchandising decisions. K-means assists managers in optimising product assortments, resource allocation, and regional marketing campaigns by identifying store-level trends.
6. A Purchase-Based Market Segmentation Methodology
 https://www.sciencedirect.com/science/article/abs/pii/S0957417404000132
Tsai and Chiu (2004) suggested a customer segmentation model based on purchasing behavior, using data-mining and clustering techniques such as K-means to group customers with similar purchase habits. Their method allowed businesses to create focused marketing programs and tailored product offers for each category. This study demonstrates how K-means transforms transactional data into strategic insights, allowing firms to build data-driven promotions and increase customer lifetime value.
 
 



Lavanya Guntupalli:
 1. Applying K-Means Clustering for User Profiling in Retail (Department Store Case Study)
 https://www.ijarsct.co.in/Paper19084.pdf
This study used K-means clustering on retail transaction and demographic data to group customers based on their expenditure and visit frequency. It discovered various consumer segments, such as regular high-spenders and occasional low-spenders, allowing the business to tailor marketing efforts, loyalty programs, and product suggestions. K-means was effective because it revealed hidden trends in massive datasets, allowing managers to focus promotions, optimise inventory for each section, and enhance retention techniques.
2. Application of K-Means for Customer Segmentation in Grocery Stores (Kenya)
 https://ijstm.inarah.co.id/index.php/ijstm/article/view/1024


This study employed K-means clustering to examine supermarket transaction data, categorizing customers based on demographics, buy frequency, and spending habits. The findings identified distinct groupings, such as high-spending young individuals and low-spending senior clients. The study found that K-means assists grocery stores in tailoring promotions, managing inventory more effectively, and identifying lucrative consumer segments for loyalty schemes, eventually enhancing customer happiness and profit margins.
3. Used Car Customer Segmentation Using K-Means (Caroline.id Case Study)
 https://www.researchgate.net/publication/379443160_Used_Car_Customer_Segmentation_Using_K-Means_Clustering_Model_With_SPSS_Program_Case_Study_CarolineId
In this case study, used-vehicle sales data were clustered using K-means to determine buyer characteristics based on age, gender, desired car type, and payment method. The technique identified two major purchasing groups, namely millennial men aged 31 to 40 who preferred automatic vehicles. K-means assisted the organization in developing targeted advertising, adjusting inventory mix, and personalizing sales methods, proving useful for businesses seeking data-driven marketing and increased consumer interaction.

4. Dynamic Pricing and Inventory Control for Multiple Products (Ríos & Vera 2023)
 https://www.sciencedirect.com/science/article/abs/pii/S036083522300089X
Although this research focuses on dynamic pricing and inventory optimization, K-means clustering may be used to improve the framework by categorizing items or consumers based on demand patterns, price sensitivity, and purchase frequency. By categorizing comparable demand habits, merchants may use personalized pricing tactics, better inventory planning, and more precisely estimate sales. This combination of K-means and optimization approaches enables more informed decision-making and improved resource allocation in retail operations.

5. Systematic Literature Review: Customer Segmentation in Retail (2019–2024)
 https://journal-iam.com/index.php/technovasia/article/view/11
This systematic evaluation examined recent customer segmentation approaches and identified K-means as one of the most successful tools for grouping retail consumers using Recency-Frequency-Monetary (RFM) characteristics. The research highlights integrating K-means with association-rule mining (e.g., FP-Growth) to identify not just consumer groups but also their product preferences. It demonstrates how K-means may assist marketing teams give targeted promotions, anticipate purchase habits, and create cluster-specific loyalty programs.

6. Implementation of K-Means Clustering for Inventory Projection (Retail Glassware Industry)
 https://www.researchgate.net/publication/381040426_Implementation_of_K-Means_Clustering_for_Inventory_Projection
K-means were employed in this study to cluster products based on demand volume and fluctuation, categorizing them as fast-, medium-, or slow-moving. The technique simplified inventory management by assisting managers in forecasting restocking requirements and avoiding overstock or shortage issues. The article revealed that K-means is a useful tool for data-driven inventory planning, increasing operational efficiency, and promoting lean retail management.


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)



library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)



df <- read_excel("Online Retail.xlsx")



glimpse(df)

# Clean data

df_clean <- df %>%
  filter(!is.na(Description), !is.na(CustomerID)) %>%     # Remove missing Description/CustomerID
  filter(Quantity > 0, UnitPrice > 0) %>%                 # Remove invalid sales
  mutate(
    InvoiceDate = as.POSIXct(InvoiceDate),
    TotalAmount = Quantity * UnitPrice,                   # Create TotalAmount column
    Month = floor_date(InvoiceDate, "month")              # Create Month column
  )
```

```{r}
#Top-selling products

top_products <- df_clean %>%
  group_by(Description) %>%
  summarise(TotalSales = sum(TotalAmount, na.rm = TRUE)) %>%
  arrange(desc(TotalSales)) %>%
  head(10)

ggplot(top_products, aes(x = reorder(Description, TotalSales), y = TotalSales)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Selling Products", x = "Product", y = "Total Sales")

```

```{r}
# Monthly sales trend

df_clean <- df_clean %>%
  mutate(Month = format(InvoiceDate, "%Y-%m"))

monthly_sales <- df_clean %>%
  group_by(Month) %>%
  summarise(TotalRevenue = sum(TotalAmount))

ggplot(monthly_sales, aes(x = Month, y = TotalRevenue, group = 1)) +
  geom_line(color = "darkgreen") +
  geom_point() +
  labs(title = "Monthly Sales Trend", x = "Month", y = "Revenue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Top countries by revenue

top_countries <- df_clean %>%
  group_by(Country) %>%
  summarise(TotalRevenue = sum(TotalAmount)) %>%
  arrange(desc(TotalRevenue)) %>%
  head(10)

ggplot(top_countries, aes(x = reorder(Country, TotalRevenue), y = TotalRevenue)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 Countries by Revenue", x = "Country", y = "Revenue")

```

```{r}
# Customer purchase distribution

customer_sales <- df_clean %>%
  group_by(CustomerID) %>%
  summarise(TotalSpent = sum(TotalAmount))

ggplot(customer_sales, aes(x = TotalSpent)) +
  geom_histogram(bins = 30, fill = "purple", color = "white") +
  labs(title = "Distribution of Customer Spending", x = "Total Spent", y = "Count")

```

```{python}
import pandas as pd
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
