---
title: "Applying K-Means Clustering in retail industry"

author: "Charan Gullipalli, Lavanya Guntupalli, Arun Pandian (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    toc: true
    toc-location: right
    toc-depth: 3
    toc-expand: true
    toc-title: "Table of Contents"
    toc-sidebar: true
    number-sections: true
    theme: flatly   
    df-print: kable
    fig-align: center
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [Presentation](slides.html){target="_blank"}

# Introduction

In today's data-driven corporate world, knowing customer behavior and
properly controlling inventory are crucial for organizational success.
Data mining and clustering techniques, such as K-Means clustering, are
crucial decision-making tools in retail, logistics, and marketing due to
increasing access to transactional, demographic, and behavioral data.
Recent studies show that algorithm selection and data pretreatment have
a direct impact on the quality of clustering insights
(@prasetiya2025review).

K-Means clustering is a machine learning technique that splits data into
distinct clusters based on similarities. This allows firms to uncover
patterns that may not be obvious through traditional analysis. Research
shows that it can significantly improve consumer segmentation, inventory
forecasting, and marketing optimization across industries. A weighted
K-Means model yielded more meaningful and stable customer clusters than
the usual model (@omol2024kenya). A review by @prasetiya2025 concluded
that K-Means is still useful in operational contexts due to its
interpretability and rapidity, even though more complex models may
provide higher precision (source).

K-Means has helped small enterprises with minimal resources in emerging
countries. Clustering transactional and demographic data helps grocery
shops detect patterns in consumer shopping behavior
(@sitorus2023inventory). Research indicates that clustering can improve
targeted marketing, loyalty programs, and stock management
(@bala2012forecast). These studies show that K-Means-based segmentation
can help small merchants coordinate supply, demand, and promotional
strategies.

Clustering methods, such as weighted variants (@omol2024kenya) and
Global and MiniBatch K-Means, have increased computing efficiency and
accuracy, addressing concerns like random initialization and scalability
on big datasets. Many research lack clear preprocessing choices, fail to
evaluate cluster quality, or use proprietary datasets that are difficult
to replicate.

This study uses K-Means clustering to resolve gaps in an open government
retail dataset and includes comprehensive documentation.

-   Steps for data cleaning and feature engineering

-   Determine the ideal number of clusters based on statistical and
    managerial criteria.

-   Validate the clusters using internal evaluation indices.

K-Means clustering is more than just a statistical technique; it's a
strategic tool for data-driven decisions. Identifying behavioral
patterns in huge datasets can help organizations improve customer
understanding, optimize resource allocation, and estimate demand more
precisely. Integrating K-Means into corporate analytics frameworks is
vital for operational excellence and competitiveness in the digital age.

## Literature Review

K-Means clustering is commonly used in retail to enhance consumer
segmentation and inventory forecasting. From 2019-2024,
@prasetiya2025review found that integrating K-Means with association
rule learning, such as FP-Growth, can help merchants better analyze
customer behavior and generate targeted marketing strategies. The review
demonstrates that K-Means is reliable on big datasets and may be used to
classify products based on sales speed, including fast-moving and
slow-moving commodities.

@omol2024 Kenya used K-Means clustering to analyze transactional and
demographic data from food outlets. Their study found that age and
income significantly impact buying behavior. This information can help
retailers better manage inventory and target marketing activities.

@bala2012forecast proposed a forecasting approach that combines
clustering and demand prediction. This technique enhanced inventory
accuracy, decreased extra stock, and minimized shortages, demonstrating
how clustering can enhance supply chain efficiency.
@sitorus2023inventory used K-Means to categorize products by demand at a
retail store, resulting in three categories: fast, moderate, and slow
sellers. This helped to optimize stock levels and lower costs.

These studies show that K-Means clustering can improve retailers'
marketing precision and inventory control by discovering client trends
and correlating them to product demand. Clustering results can help
match promotional campaigns with inventory decisions, resulting in a
data-driven retail management system (@prasetiya2025review;
@omol2024kenya; @bala2012forecast; @sitorus2023inventory).

# Methods

K-Means is an unsupervised machine learning algorithm that groups
related data points into distinct clusters. The algorithm divides data
into K groups based on value similarity. Each group has a centroid,
where all items are located in the center.

The K-Means approach separates a set of observations (x1,x2,…,xn) into K
distinct clusters (C1,C2,…,CK) by minimizing the Within-Cluster Sum of
Squares (WCSS):

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$

where uk is the centroid (mean vector) of cluster Ck and \|\|.\|\|
denotes the Euclidean norm.

The goal is to create clusters that are both compact (low variance
inside) and distinct (big separation between centers).

This project use the K-Means method to categorize items or suppliers
based on their performance and sales behavior. The primary goal is to
identify patterns, such as which things sell quickly, moderately, or
slowly. Creating clusters helps us identify sales trends and make better
inventory and management decisions.

This helps with:

-   Determine whether things move quickly, moderately, or slowly.

-   Analyze sales trends and supplier performance.

-   Improve decision-making for planning and replenishing.

-   Break down enormous amounts of data into relevant, understandable
    groups.

In simple terms: This project uses K-Means clustering to automatically
categorize commodities or suppliers based on their behavior.
Transforming complex data into clear insights enhances understanding of
trends, decision-making, and inventory management.

## Determining the Optimal Number of Clusters

Choosing an appropriate number of clusters (𝐾) is critical for effective segmentation.  Three complimentary approaches were utilized:

1.  **Elbow Method**: The Within-Cluster Sum of Squares (WCSS) was compared to increasing values of 𝐾.  The optimal value is indicated by a dramatic fall in the pace of improvement ("the elbow"). (@MacQueen1967)

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$

2.  **Silhouette Coefficient**: This metric assesses how well each observation fits in its allocated cluster compared to nearby clusters, with values ranging from -1 (poor) to +1 (well-separated). (@Rousseeuw1987)

$$
s(i) = \frac{b(i) - a(i)}{\max(a(i), b(i))}
$$

3.  **Calinski–Harabasz Index**: The Calinski-Harabasz (CH) index compares between and within cluster dispersion. A higher CH index in 𝐾clusters suggests better defined clusters.(@Calinski1974)
     
We compared 𝐾 values ranging from 2 to 6.

- The elbow plot of WCSS.

- The average Silhouette score,

- The Calinski-Harabasz score.

**Practical Interpretation and Final Choice** The Elbow and Silhouette analyses indicated that 𝐾=3 strikes a fair balance between model complexity and cluster separation.  Furthermore, 𝐾=3 is relevant for managers as it aligns with the traditional inventory classification system.

- Fast moving items,

- Moderately moving items,

- Slow-moving items

 The ideal number of clusters for this investigation was determined to be 𝐾 = 3.

## K-Means Clustering Process

After selecting =3, the K-Means technique was used to analyze the standardized feature space (TOTAL_SALES, RETAIL SALES, WAREHOUSE SALES).

 The main algorithmic steps (@Lloyd1982) are:
 
1.**Initialization:**

-  Randomly set 𝐾 = 3 centroids in the feature space, or

-  Use K-Means initialization for better beginning positions.
 
2.**Assignment step:** For each item (x_i), compute the Euclidean
distance to each centroid (u_k):

$$
d(x_i, \mu_k) = \| x_i - \mu_k \|_2
$$

Assign (x_i) to the cluster with the closest centroid.

3.**Update step:** -For each cluster Ck, recompute the centroid as the
mean of its members:

$$
\mu_k = \frac{1}{|C_k|} \sum_{x_i \in C_k} x_i
$$

4.**Convergence:**

-   Repeat the assignment and update stages until the centroids stabilize (changes below a specified threshold) or the maximum number of iterations is reached.  Our investigations showed that convergence often happened after a few repetitions.


## Cluster Validation and Profiling

After fitting the K-Means model, the cluster quality was evaluated using:

- The average Silhouette score, which measures coherence and separation;

- The Calinski-Harabasz index compares variation between and within clusters.
 
The indices give quantitative validation, but clusters were also profiled qualitatively to connect them to commercial principles.  For each cluster, we analyzed:

- The average and median of total sales, retail sales, and warehouse sales;

- Distribution of item types, including beer, wine, liquor, kegs, and non-alcoholic beverages.

- The contribution of major suppliers

- Using monthly summary to identify temporal patterns.

Clusters were classified as Fast, Moderate, or Slow-Moving based on their profiles.
 
# Analysis and Results

## Data Exploration and Visualization

**Data Sources and Collection:**

The Montgomery County Maryland Open Data Portal (@MontgomeryCountyWarehouseRetail2025) provided the dataset for this investigation.

The Department of Liquor Control collected complete monthly statistics on alcohol sales from 2017 to 2020 from retail and warehouse locations. Each record includes information regarding:

-   Supplier name (vendor providing the product)

-   item kind (beer, wine, spirits, etc.) and item code

-   Retail sales, retail transfers, and warehouse sales are a few such
    sales channels.

-   Temporal features: month and year

The dataset is constantly updated and released under a Public Domain License, promoting transparency in local government processes and facilitating replicable business analytics research.

## Exploratory Analysis

Before using K-Means, the dataset was cleaned and summarized to analyze sales trends, supplier performance, and product movement.  Using RStudio and tools like tidyverse, ggplot2, and lubridate, we created descriptive statistics and exploratory visualizations.

 An example of cleansed data, aggregated at the item level, includes variables such as:
 
-   YEAR, MONTH

-   SUPPLIER, ITEM CODE, ITEM DESCRIPTION, ITEM TYPE

-   RETAIL SALES, RETAIL TRANSFERS, WAREHOUSE SALES

-   TOTAL_SALES

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)


# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)



data <- read_excel("Retail_Sales.xlsx")







# Remove duplicate rows
data <- distinct(data)

# Handle missing values (replace or remove)
data <- data %>%
  mutate(
    SUPPLIER = ifelse(is.na(SUPPLIER), "Unknown Supplier", SUPPLIER),
    `ITEM TYPE` = ifelse(is.na(`ITEM TYPE`), "Unknown Type", `ITEM TYPE`),
    `RETAIL SALES` = ifelse(is.na(`RETAIL SALES`), 0, `RETAIL SALES`),
    `RETAIL TRANSFERS` = ifelse(is.na(`RETAIL TRANSFERS`), 0, `RETAIL TRANSFERS`),
    `WAREHOUSE SALES` = ifelse(is.na(`WAREHOUSE SALES`), 0, `WAREHOUSE SALES`)
  )

# Create a Total Sales column
data <- data %>%
  mutate(TOTAL_SALES = `RETAIL SALES` + `RETAIL TRANSFERS` + `WAREHOUSE SALES`)

# Combine YEAR and MONTH into a proper Date column
data <- data %>%
  mutate(Date = make_date(YEAR, MONTH, 1))


```


**Total Sales Trend Over Time**
```{r}
# Total Sales Trend Over Time

library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)

# Ensure the Date column is in date format
data <- data %>%
  mutate(Date = as.Date(Date))

# Aggregate sales by date (if multiple transactions per day)
daily_sales <- data %>%
  group_by(Date) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

# Line chart for total sales trend
ggplot(daily_sales, aes(x = Date, y = Total_Sales)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "orange", size = 1) +
  labs(
    title = "Total Sales Trend Over Time",
    x = "Date of Transaction",
    y = "Total Daily Sales (in USD)"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

**Insights:** This image depicts the daily total sales trend over time.  It shows daily sales fluctuations, including peaks during high demand and dips during slack days.  The trend line identifies overall performance changes and seasonal differences in sales.


**Top 10 Suppliers by Total Sales**
```{r}
# Top 10 Suppliers by Total Sales

top_suppliers <- data %>%
  group_by(SUPPLIER) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

ggplot(top_suppliers, aes(x = reorder(SUPPLIER, Total_Sales), y = Total_Sales, fill = SUPPLIER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Suppliers by Total Sales", x = "Supplier", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**Sales analysis shows that a small number of suppliers account for the majority of total income.  Crown Imports, Miller Brewing Company, and Anheuser Busch Inc. have the highest sales, indicating a concentrated supply base where major vendors own the majority of the market.  Maintaining strong supplier connections and accurate demand forecasting is crucial for these significant contributors.


**Sales by Item Type**
```{r}
# Sales by Item Type

ggplot(data, aes(x = `ITEM TYPE`, y = TOTAL_SALES, fill = `ITEM TYPE`)) +
  geom_bar(stat = "summary", fun = "sum") +
  labs(title = "Total Sales by Item Type", x = "Item Type", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**Beer dominates the retail market, accounting for about $8 million in sales and exceeding all other categories based on item distribution.  Liquor and wine fall into second and third, indicating moderate but continuous demand.  Alcoholic beverages dominate the county's retail industry, with dunnage, kegs, and non-alcoholic items accounting for only a small portion of total income.

- Beer's dominance indicates a fast-moving inventory sector due to regular use and strong brand loyalty.  Goods classified as "Kegs" and "Non-Alcoholic" move slowly and require targeted marketing to increase sales.

**Comparison: Retail vs Warehouse Sales**
```{r}
# Comparison: Retail vs Warehouse Sales

sales_compare <- data %>%
  summarise(
    Retail_Sales = sum(`RETAIL SALES`, na.rm = TRUE),
    Warehouse_Sales = sum(`WAREHOUSE SALES`, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Category", values_to = "Sales")

ggplot(sales_compare, aes(x = Category, y = Sales, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Retail vs Warehouse Sales", x = "", y = "Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:** Warehouse sales are nearly three times higher than retail sales, according to the data.  Merchandise is typically distributed in bulk through centralized channels rather than being sold directly to clients.  While this structure increases productivity, it requires careful inventory balance to avoid overstocking.

**Monthly Sales Heatmap**
```{r}

# Monthly Sales Heatmap
library(viridis)

monthly_sales <- data %>%
  group_by(YEAR, MONTH) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

ggplot(monthly_sales, aes(x = factor(MONTH), y = factor(YEAR), fill = Total_Sales)) +
  geom_tile() +
  scale_fill_viridis(option = "magma", direction = -1, labels = comma) +
  labs(title = "Monthly Sales Heatmap", x = "Month", y = "Year", fill = "Total Sales") +
  theme_minimal()

```

**Insights:**Sales activity often peaks between December and the middle of the year, driven by seasonal and holiday demand.  Early months show reduced sales, while anomalies in 2020 indicate potential external interruptions such the COVID-19 pandemic.  Identifying seasonal changes helps plan stock levels more effectively.

## Visualization-Based Evidence

1.**Top Suppliers by Total Sales:**A bar chart of total sales by source revealed that Crown Imports, Miller Brewing Company, and Anheuser Busch Inc. dominate revenue.  The company's primary items are concentrated in Cluster 1 (Fast-Moving), resulting in high overall sales and strong warehouse movement.  Smaller suppliers typically have more products in Clusters 2 (slow-moving) or 3 (moderate-moving).


2.**Total Sales by Item Type:**Beer is the most popular item in overall sales, followed by spirits and wine.  Cluster 1 has a significant concentration of fast-moving products, particularly in high-volume categories such as beer and liquor.  Cluster 3 (Moderate-Moving) features wine and mid-tier products with consistent but lower sales.  Kegs, dunnage, and non-alcoholic commodities fall under Cluster 2 (Slow-Moving) due to their slow or irregular turnover.


3.**Retail vs Warehouse Sales:** Warehouse sales ($7.8M) are significantly higher than retail sales ($2.1M).  Cluster 1 (Fast-Moving) items have substantial warehouse sales and retail movement, indicating they are stored in bulk to meet ongoing demand.  Items in Cluster 3 (Moderate-Moving) represent mid-level warehouse and retail sales.  Cluster 2 (Slow-Moving) goods have low values in both channels and are rarely recorded in sales statistics.

4.**Heatmap of Monthly Sales:** A monthly heatmap of sales activity indicates that fast-moving items (Cluster 1) experience seasonal peaks, particularly in summer and December, but still have large baseline sales throughout the year.  Moderate-moving goods (Cluster 3) exhibit consistent patterns throughout months.  Slow-moving products (Cluster 2) often have minimal or no sales, with rare increases.  COVID-19 may have contributed to the unpredictable behavior in 2020.


These visual patterns complement the mathematical cluster profiles and support the classification of Clusters 1, 2, and 3 as fast, slow, and moderate moving groups.

# Modeling and Results

## Modeling and Data Preparation

To develop the K-Means model, the dataset was prepared to ensure accurate and consistent item representation.  Duplicate records were deleted, missing values were addressed, and sales from three channels (retail, warehouse, and retail transfers) were integrated into a single TOTAL_SALES metric to accurately reflect item performance.
These cleansed and condensed variables laid the groundwork for clustering.

The Elbow and Silhouette Methods were used to determine the optimal number of clusters. Both methodologies consistently showed K = 3, which corresponds to standard retail inventory classifications (Fast-Moving, Moderate-Moving, and Slow-Moving). Using K, the dataset was ready for clustering based on standardized sales variables.

## Data Preprocessing

The Montgomery County retail dataset included monthly sales records for all suppliers, items, and sales channels. To prepare the data for K-Means, multiple preprocessing processes were performed:

-   **Aggregation**: Monthly records were combined into one row per supplier-item combination, including Retail Sales, Retail Transfers, and Warehouse Sales.

-   **Feature Creation**: A new variable, TOTAL_SALES, was created to reflect all sales activity.

-   **Missing Values**: Missing numeric sales values were substituted with zeros when applicable.

-   **Reduplication**: Duplicate entries have been removed to prevent excess specific items.

-   **Filtering**: Items with no sales activity were removed from modeling.

-   **Standardization**: Clustering variables (TOTAL_SALES, RETAIL_SALES, and WAREHOUSE_SALES) were normalized to prevent scale gaps from affecting distance calculations.

The preparation methods ensured a clean, consolidated dataset suited for cluster creation.

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(cluster)    # for silhouette()
library(factoextra) # for convenient plots (optional)
library(scales)   

# Read Excel file
retail_raw <- read_excel("Retail_Sales.xlsx")

# Aggregate to item level: one row per supplier-item
items <- retail_raw %>%
  group_by(SUPPLIER, `ITEM CODE`, `ITEM DESCRIPTION`, `ITEM TYPE`) %>%
  summarise(
    RETAIL_SALES     = sum(`RETAIL SALES`, na.rm = TRUE),
    RETAIL_TRANSFERS = sum(`RETAIL TRANSFERS`, na.rm = TRUE),
    WAREHOUSE_SALES  = sum(`WAREHOUSE SALES`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    TOTAL_SALES = RETAIL_SALES + RETAIL_TRANSFERS + WAREHOUSE_SALES
  ) %>%
  filter(TOTAL_SALES > 0)  # keep only items that actually sold something

# Numeric features for clustering
features <- items %>%
  select(TOTAL_SALES, RETAIL_SALES, WAREHOUSE_SALES)

# Scale features
features_scaled <- scale(features)
```

## Elbow Method Code

The Elbow Method was used to calculate the Within-Cluster Sum of Squares (WCSS) for 𝐾 values ranging from 2 to 8. The Elbow plot illustrates how WCSS reduces with more clusters. A sharp drop followed by a flattening of the curve, known as a "elbow," shows a strong balance of simplicity and precision.

```{r}
set.seed(123)

k_values <- 2:8

wcss <- map_dbl(k_values, function(k) {
  kmeans(features_scaled, centers = k, nstart = 10)$tot.withinss
})

elbow_df <- tibble(
  k = k_values,
  WCSS = wcss
)

ggplot(elbow_df, aes(x = k, y = WCSS)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Elbow Plot for Choosing K",
    x = "Number of Clusters (K)",
    y = "Total Within-Cluster Sum of Squares (WCSS)"
  ) +
  theme_minimal()
```

The WCSS curve exhibited a clear bend at 𝐾=3.  Beyond three clusters, the drop in WCSS was small, indicating that adding more clusters would not considerably enhance separation.  The results suggest that segmenting retail items into three clusters is an effective and efficient approach.

Using WCSS patterns, 𝐾=3 creates a meaningful structure with minimal unneeded complexity.

## Silhouette Method

To validate the Elbow Method, a Silhouette analysis was performed for 𝐾 = 3.  The Silhouette score assesses how closely each item aligns with its own cluster compared to others.  Clusters with values near +1 show strong cohesiveness and distinct separation.

```{r}
set.seed(123)

# Fit K-Means with K = 3
k <- 3
k3_model <- kmeans(features_scaled, centers = k, nstart = 25)

# Silhouette object
sil_obj <- silhouette(k3_model$cluster, dist(features_scaled))

# Average silhouette width (you can quote this number in your text)
mean_silhouette <- mean(sil_obj[, "sil_width"])
mean_silhouette

# Plot silhouette for K = 3
fviz_silhouette(sil_obj) +
  labs(title = "Silhouette Plot for K = 3")
```

The average Silhouette score was 0.97, indicating outstanding clustering performance.  Cluster 2 had a Silhouette width of 0.98, indicating tight grouping and significant separation, perhaps due to its high number of low-sales goods.  Clusters 1 and 3 had lower Silhouette values (0.19 and 0.09), consistent with smaller high-value and mid-value groups located further away from the main slow-moving cluster.

The high overall Silhouette score indicates that 𝐾=3 creates well-defined and meaningful clusters.

## Cluster Profiling(Sizes + Centroids + Summary)

After clustering, the features of each cluster were analyzed using their sizes, centroid values (scaled units), and summary statistics (original units).  This profiling stage translates numerical cluster assignments into relevant business categories.

```{r}
# Add cluster labels back to item data
items_clustered <- items %>%
  mutate(cluster = factor(k3_model$cluster))
```

```{r}
# Cluster sizes (how many items in each cluster)
cluster_sizes <- items_clustered %>%
  count(cluster)

cluster_sizes
```

The cluster size results showed a very uneven distribution:

-   Cluster 1: 17 items

-   Cluster 2: 37,667 items

-   Cluster 3: 240 items

```{r}
# Centroids on scaled features (for technical report)
cluster_centroids_scaled <- as_tibble(k3_model$centers) %>%
  mutate(cluster = factor(1:3))

cluster_centroids_scaled
```

```{r}

# Summary stats in original units (mean and median)
cluster_summary <- items_clustered %>%
  group_by(cluster) %>%
  summarise(
    mean_total_sales   = mean(TOTAL_SALES, na.rm = TRUE),
    median_total_sales = median(TOTAL_SALES, na.rm = TRUE),
    mean_retail_sales  = mean(RETAIL_SALES, na.rm = TRUE),
    mean_wh_sales      = mean(WAREHOUSE_SALES, na.rm = TRUE),
    .groups = "drop"
  )

cluster_summary
```

The centroids and averages revealed that:

-  Cluster 1 has the most overall sales (about 139,000 on average), making it the Fastest Moving cluster.

-   Cluster 2 has low total sales (about 155), indicating slow moving products.

-  Cluster 3 has moderate overall sales (about 16,695), making it the Moderate-Moving cluster.

This three-group structure aligns with common retail inventory movement patterns and managerial stock classification frameworks.  K-Means accurately identified high-demand items from mid-level and slow-moving products, as confirmed by the profile.

## Cluster Visualization: Scatter Plot

```{r}
ggplot(items_clustered,
       aes(x = WAREHOUSE_SALES,
           y = TOTAL_SALES,
           color = cluster)) +
  geom_point(alpha = 0.5, size = 1.1) +
  
  # X-axis: log10, start at 100, no gap
  scale_x_log10(
    labels = scales::comma,
    limits = c(100, NA),
    expand = c(0, 0)   # <-- removes the gap at the axis
  ) +
  
  # Y-axis: log10, start at 100, no gap
  scale_y_log10(
    labels = scales::comma,
    limits = c(100, NA),
    expand = c(0, 0)   # <-- removes the gap at the axis
  ) +
  
  labs(
    title = "K-Means Clusters",
    x = "Warehouse Sales",
    y = "Total Sales",
    color = "Cluster"
  ) +
  theme_minimal()
```

**Interpretation**

-   Cluster 1: points in high-sales region → Fast-Moving.

-   Cluster 2: dense cloud near origin → Slow-Moving.

-   Cluster 3: middle zone → Moderate-Moving.

## Cluster Formation and Categorization

Using K-Means with 𝐾=3, the dataset was separated into three clusters based on Total Sales, Retail Sales, and Warehouse Sales.

Using the cluster summary results:

-   Cluster 1 has only 17 goods yet generates an average total sales of 138,991, including significant warehouse sales.

-   Cluster 2 has 37,667 products, but has a low average total sales of 155, as well as low retail and warehouse sales.

-   Cluster 3 comprises 240 items, with an average total sales of approximately 16,695, and intermediate warehouse and retail values.

Based on these patterns, the clusters were interpreted as:

-   Cluster 1 – Fast-moving items This cluster contains items with high total and warehouse sales. These are the "core" products that generate significant revenue due to their high volume sales.
High-demand items require frequent restocking and close monitoring to prevent stock-outs.

-   Cluster 2 – Slow Moving Items  This cluster includes the bulk of goods yet has low average sales.  Some of these products have limited or irregular demand, and may be considered niche, specialist, or infrequently purchased.  They feature lower retail and warehousing activity compared to other clusters.

-   Cluster 3 – Moderately Moving Items  This cluster includes items that fall somewhere in the middle.  Their sales are significantly greater than slow movers, but lower than quick movers in Cluster 1.  They demonstrate consistent and predictable performance throughout time.  These products are typically wines and mid-tier spirits with a consistent but not high turnover.

This structure aligns with the conventional inventory classification of rapid, moderate, and slow-moving products, indicating that K-Means accurately reflects sales behavior.

## Cluster Insights

**Fast-Moving Items (Cluster 1)** Fast-moving items are a small group of products that generate high total and warehouse sales.  Key vendors such as Crown Imports and Miller Brewing supply major beer brands and high-quality liquor.  Effective management of these sources is crucial since they contribute significantly to overall revenue.

-   To avoid stockouts, they require constant replenishing.

-   Forecasting inaccuracies can significantly effect customer satisfaction and revenue.

-   These products should be prioritized in storage, shelf space, and monitoring.

**Slow-Moving Items (Cluster 2)** Slow-moving products comprise the largest cluster, although each sells in limited quantities.  This category includes specialty, seasonal, or low-demand products, such as kegs and non-alcoholic beverages.

-   They have a larger risk of overstocking and inventory holding costs.

-  Managers can consider lowering order numbers, revising assortment, or offering promotions and discounts to move slow stock.

-   Long-term slow movers may be discontinued if they do not align with strategic aims.

**Moderate-Moving Items (Cluster 3)** Moderate-moving objects fall between the other two groupings.  They have average total sales and consistent, predictable demand. This category includes many mid-tier wines and spirits.

-   They generate consistent revenue and contribute to a diverse product offering.

-   They benefit from a well-balanced safety stock, which is neither too aggressive nor too restrictive.

-   Regular performance reviews can determine whether goods should be promoted to more active categories or reduced if demand decreases.

Overall, The three clusters provide a clear structure for the product portfolio.  Managers might prioritize protecting the supply of fast-moving commodities, stabilizing moderate items, and reducing costs for slow-moving items.

# Conclusion

## Key Findings

-  K-Means clustering classified products into three categories: Fast-Moving, Moderate-Moving, and Slow-Moving items.

-   Beer and liquor brands from vendors like Crown Imports and Miller Brewing had high and consistent sales.  These goods must be supplied often.

-  Moderate-moving commodities, like wine, had average and consistent sales.  They are stable products that require balanced inventories.

-   Slow-moving items were typically niche or seasonal, with low or inconsistent sales.  To prevent overstocking and storage expenses, these products must be carefully monitored.

-   Sales increased during holiday seasons (e.g., summer and December), indicating that demand varies with time and events.

## Results Summary

The clustering results helped group inventory based on sales behavior
and stock movement.

This makes it easier for managers to:

-   Focus on fast-selling items to keep shelves full,

-   Maintain enough stock of moderate-selling items, and

-   Reduce extra inventory of slow-selling items.

K-Means clustering improves inventory planning, reduces waste, and enables quick response to client demand. Overall, it improves inventory control, forecasting, and retail efficiency.

## Limitations and Future Work

**Limitations**

**Cluster Size Imbalance:**One drawback is that Cluster 2 is significantly larger than Clusters 1 and 3.  This reflects the fact that most retail items move slowly, but it also indicates that the clustering structure is severely uneven.  This could affect how well the smaller high-value clusters are represented.

**K-Means Assumptions:**K-Means assumes spherical, uniformly sized clusters and is sensitive to initial centroid location and outliers.  Given the skewed distribution of retail sales data, these assumptions may not always be valid.

**Limited Feature Set:**The analysis only considers sales-related factors (total sales, retail sales, and warehouse sales).  Important business parameters, such as profit margin, price elasticity, promotional activity, consumer demographics, and item lifetime, were not considered and could have an impact on cluster formation.

**Future Work**

**Test Alternative Clustering Algorithms:**Investigate models that can handle skewed and imbalanced data, such as DBSCAN, Gaussian Mixture Models, and hierarchical clustering, to see if a more balanced segmentation emerges.

**Add more business features:**Use margin, profitability, seasonality indications, promotions, or RFM-style customer KPIs to create clusters that reflect both sales volume and business value.

**Try Weighted or Hybrid models:**Consider weighted K-Means, MiniBatch K-Means, or hybrid models to reduce sensitivity to extreme values and improve resilience.

**Extend to customer-level clustering:**Using the same dataset, cluster consumer behavior (frequency, recency, spend, brand loyalty) to supplement product-level segmentation.

