---
title: "Applying K-Means Clustering in retail industry"

author: "Charan Gullipalli, Lavanya Guntupalli, Arun Pandian (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    toc: true
    toc-location: right
    toc-depth: 3
    toc-expand: true
    toc-title: "Table of Contents"
    toc-sidebar: true
    number-sections: true
    theme: flatly   
    df-print: kable
    fig-align: center
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

# Introduction

In today's data-driven corporate environment, understanding customer
behavior and effectively managing inventories are key to organizational
success. With the increased availability of transactional, demographic,
and behavioral data, data mining and clustering techniques—particularly
K-Means clustering—have become critical decision-making tools in retail,
logistics, and marketing. Recent studies emphasize that algorithm
selection and data preprocessing directly affect the quality of insights
derived from clustering (@prasetiya2025review).

K-Means clustering is a machine learning technique that divides data
into distinct clusters based on similarities, allowing businesses to
identify patterns that would not be visible using traditional analysis.
Research has demonstrated that it can greatly enhance consumer
segmentation, inventory forecasting, and marketing optimization across
various industries. For example, a weighted K-Means model produced more
meaningful and stable customer clusters compared to the standard model
(@omol2024kenya). Similarly, comparative work on the same dataset found
that while more complex models can sometimes yield higher precision,
K-Means remains valuable for its interpretability and speed in
operational settings (@prasetiya2025review).

In emerging markets, K-Means has also proven beneficial for small and
resource-limited businesses. Clustering transactional and demographic
data enables grocery stores to identify clear patterns of consumer
purchasing behavior (@sitorus2023inventory). Likewise, studies show that
clustering supports targeted marketing, loyalty programs, and more
efficient stock management (@bala2012forecast). Collectively, these
works underscore that even small or localized retailers can leverage
K-Means-based segmentation to align supply, demand, and promotional
strategy.

Beyond retail, advances in clustering algorithms—including weighted
variants (@omol2024kenya) as well as Global and MiniBatch K-Means—have
improved computational efficiency and accuracy, reducing issues like
random initialization and scalability on large datasets. However, many
studies still under-specify preprocessing choices, do not rigorously
validate cluster quality, or rely on proprietary datasets that are not
easily reproducible.

This study addresses these gaps by applying K-Means clustering to an
open, government retail dataset and by explicitly documenting:

-   The data cleaning and feature engineering steps,

-   The selection of the optimal number of clusters using both
    statistical and managerial criteria, and

-   The validation of the resulting clusters using internal evaluation
    indices.

In summary, K-Means clustering is more than a statistical technique; it
is a strategic enabler for data-informed decision-making. By uncovering
relevant behavioral patterns in large datasets, firms can deepen
customer understanding, optimize resource allocation, and forecast
demand more accurately. Integrating K-Means into corporate analytics
frameworks is thus a crucial step toward operational excellence and
maintaining competitiveness in the digital economy.

## Literature Review

Research shows that K-Means clustering is widely used in retail to
improve customer segmentation and inventory forecasting. According to
@prasetiya2025review, studies between 2019–2024 identified that
combining K-Means with association rule learning (like FP-Growth) helps
retailers better understand customer behavior and create focused
marketing campaigns. The review also highlights that K-Means performs
reliably on large datasets and can be applied to classify products by
sales speed — such as fast-moving and slow-moving goods.

@omol2024kenya applied K-Means clustering to grocery stores in Kenya
using transactional and demographic data. Their study showed that age
and income strongly influence shopping behavior. Retailers could use
this information to manage inventory and target marketing efforts more
effectively.

@bala2012forecast introduced a forecasting model combining clustering
and demand prediction. This approach improved inventory accuracy,
reduced excess stock, and minimized shortages — showing that clustering
can make supply chains leaner and more efficient. Similarly,
@sitorus2023inventory used K-Means to group products by demand at a
retail store and identified three categories: fast, moderate, and slow
sellers. This helped optimize stock levels and reduce costs.

Together, these studies demonstrate that K-Means clustering helps
retailers improve both marketing precision and inventory control by
identifying customer patterns and linking them to product demand. For
example, clustering results can be used to align promotional strategies
with inventory decisions, creating an integrated, data-driven retail
management system (@prasetiya2025review; @omol2024kenya;
@bala2012forecast; @sitorus2023inventory).

# Methods

K-Means is an unsupervised machine learning method used to group similar
data points into different clusters. It works by dividing all the data
into a specific number of groups (called K) based on how close or
similar the values are to each other. Each group has a center point
called a centroid, and all the items in that group are close to that
center.

Formally, given a set of n observations x1,x2,…,xn the K-Means algorithm
partitions the data into K disjoint clusters C1,C2,…,CK by minimizing
the Within-Cluster Sum of Squares (WCSS):

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$ where uk is the centroid (mean vector) of cluster Ck and \|\|.\|\|
denotes the Euclidean norm.

The overall goal is to make each cluster as compact as possible (low
variance within clusters) and as distinct as possible from the others
(large separation between cluster centers).

In this project, the K-Means method is used to group items or suppliers
based on their performance and sales behavior. The main idea is to find
patterns — such as which products sell quickly, moderately, or slowly.
By forming these clusters, we can easily understand sales trends and
improve inventory or management decisions.

This helps to:

-   Identify fast-moving, moderate-moving, and slow-moving items.

-   Analyze sales patterns and supplier performance.

-   Support better decision-making for planning and restocking.

-   Simplify large data into meaningful, easy-to-understand groups.

In Simple Terms: This project uses K-Means clustering to automatically
divide items or suppliers into groups based on how they behave. It helps
to understand trends, improve decisions, and manage inventory more
effectively by turning complex data into clear, useful insights.

## Determining the Optimal Number of Clusters

Selecting an appropriate number of clusters 𝐾 is crucial for meaningful
segmentation. Three complementary techniques were used:

1.  **Elbow Method**: The Within-Cluster Sum of Squares (WCSS) was
    plotted against increasing values of 𝐾. The point where the rate of
    improvement sharply declines (“the elbow”) suggests an optimal
    value.(@MacQueen1967)

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$

2.  **Silhouette Coefficient**: Measures how well each observation fits
    within its assigned cluster versus neighboring clusters, ranging
    from -1 (poor) to +1 (well-separated).(@Rousseeuw1987)

$$
s(i) = \frac{b(i) - a(i)}{\max(a(i), b(i))}
$$

3.  **Calinski–Harabasz Index**: The Calinski–Harabasz(CH) index
    compares between-cluster dispersion with within-cluster dispersion.
    For 𝐾clusters, a higher CH index indicates better defined clusters.

In practice, we evaluated a range of 𝐾 values (e.g., 𝐾=2 to 𝐾=6) and
compared:

-   The Elbow plot of WCSS,

-   The average Silhouette scores, and

-   The Calinski–Harabasz scores.

**Practical Interpretation and Final Choice** Both the Elbow and
Silhouette analyses suggested that 𝐾=3 provides a good balance between
model complexity and cluster separation. In addition, 𝐾=3 is
managerially meaningful, aligning with the conventional classification
of inventory into:

-   Fast-moving items,

-   Moderate-moving items, and

-   Slow-moving items.

Therefore, 𝐾=3 was chosen as the optimal number of clusters for this
study.

## K-Means Clustering Process

Once =3 was selected, the K-Means algorithm was applied to the
standardized feature space (TOTAL_SALES,RETAIL SALES,WAREHOUSE SALES).

The main algorithmic steps(@Lloyd1982) are:

1.**Initialization:**

-   Randomly initialize 𝐾 =3 centroids in the feature space, or

-   Use a smarter method such as K-Means++ initialization to improve
    starting positions.

2.**Assignment step:** For each item (x_i), compute the Euclidean
distance to each centroid (u_k):

$$
d(x_i, \mu_k) = \| x_i - \mu_k \|_2
$$

Assign (x_i) to the cluster with the closest centroid.

3.**Update step:** -For each cluster Ck, recompute the centroid as the
mean of its members:

$$
\mu_k = \frac{1}{|C_k|} \sum_{x_i \in C_k} x_i
$$

4.**Convergence:**

-   Repeat the assignment and update steps until the centroids stabilize
    (i.e., changes fall below a chosen threshold) or a maximum number of
    iterations is reached. In our experiments, convergence typically
    occurred within a small number of iterations.

## Cluster Validation and Profiling

After fitting the K-Means model, cluster quality was assessed using:

-   The average Silhouette score, to gauge cohesion and separation;

-   The Calinski–Harabasz index, to compare between-cluster vs.
    within-cluster variance.

While these indices provide quantitative validation, clusters were also
profiled qualitatively to link them to business concepts. For each
cluster, we examined:

-   Mean and median of TOTAL_SALES, RETAIL SALES, and WAREHOUSE SALES;

-   Distribution of item types (beer, wine, liquor, kegs,
    non-alcoholic);

-   Contribution of major suppliers;

-   Temporal patterns using monthly summaries.

Clusters were then labeled as Fast-Moving, Moderate-Moving, or
Slow-Moving based on these profiles.

# Analysis and Results

## Data Exploration and Visualization

**Data Sources and Collection:**

The Montgomery County Maryland Open Data
Portal(@MontgomeryCountyWarehouseRetail2025) provided the dataset used
in the present study.

It includes comprehensive monthly records of alcohol sales from 2017 to
2020 that were gathered by the Department of Liquor Control from both
retail locations and warehouse operations. Every record contains details
regarding:

-   Supplier name (vendor providing the product)

-   item kind (beer, wine, spirits, etc.) and item code

-   Retail sales, retail transfers, and warehouse sales are a few such
    sales channels.

-   Temporal features: month and year

The dataset is updated regularly and published under a Public Domain
License, supporting transparency in local government operations and
enabling reproducible business analytics research.

## Data Preprocessing

Before applying K-Means, the dataset was cleaned, aggregated, and
transformed to ensure accurate clustering performance. The original
retail dataset contained monthly-level sales for each supplier, item
code, and sales channel (retail, transfers, warehouse). These records
were aggregated to the item level, creating a single row per
supplier-item combination with total sales, retail sales, and warehouse
sales.

Missing values were replaced with zeros where appropriate, duplicates
were removed, and a new TOTAL_SALES variable was created by combining
all three sales channels. Only items with at least some sales activity
were retained. Finally, numerical variables used for clustering were
standardized to remove scale differences between retail and warehouse
sales.

This preprocessing ensured that the clustering model was applied to
cleaned, consistent, and comparable data.

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(cluster)    # for silhouette()
library(factoextra) # for convenient plots (optional)

# Read Excel file
retail_raw <- read_excel("Retail_Sales.xlsx")

# Aggregate to item level: one row per supplier-item
items <- retail_raw %>%
  group_by(SUPPLIER, `ITEM CODE`, `ITEM DESCRIPTION`, `ITEM TYPE`) %>%
  summarise(
    RETAIL_SALES     = sum(`RETAIL SALES`, na.rm = TRUE),
    RETAIL_TRANSFERS = sum(`RETAIL TRANSFERS`, na.rm = TRUE),
    WAREHOUSE_SALES  = sum(`WAREHOUSE SALES`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    TOTAL_SALES = RETAIL_SALES + RETAIL_TRANSFERS + WAREHOUSE_SALES
  ) %>%
  filter(TOTAL_SALES > 0)  # keep only items that actually sold something

# Numeric features for clustering
features <- items %>%
  select(TOTAL_SALES, RETAIL_SALES, WAREHOUSE_SALES)

# Scale features
features_scaled <- scale(features)
```

## Elbow Method Code

The Elbow Method was used to evaluate different values of 𝐾 by
calculating the Within-Cluster Sum of Squares (WCSS) for 𝐾=2 through
𝐾=8. The Elbow plot shows how much WCSS decreases as the number of
clusters increases. A sharp drop is usually followed by a flattening of
the curve—this “elbow” indicates a good balance between simplicity and
accuracy.

```{r}
set.seed(123)

k_values <- 2:8

wcss <- map_dbl(k_values, function(k) {
  kmeans(features_scaled, centers = k, nstart = 10)$tot.withinss
})

elbow_df <- tibble(
  k = k_values,
  WCSS = wcss
)

ggplot(elbow_df, aes(x = k, y = WCSS)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Elbow Plot for Choosing K",
    x = "Number of Clusters (K)",
    y = "Total Within-Cluster Sum of Squares (WCSS)"
  ) +
  theme_minimal()
```

The WCSS curve showed a clear bend at 𝐾=3. Beyond three clusters, the
reduction in WCSS became minimal, suggesting that additional clusters
would not significantly improve separation. This result supports
choosing three clusters as an appropriate and efficient segmentation of
the retail items.

Thus, based on WCSS patterns, 𝐾=3 provides a meaningful structure with
minimal unnecessary complexity.

## Silhouette Method

To validate the Elbow Method, a Silhouette analysis was performed for
𝐾=3. The Silhouette score measures how close each item is to its own
cluster compared to other clusters. Values close to +1 indicate strong
cohesion within clusters and clear separation between them.

```{r}
set.seed(123)

# Fit K-Means with K = 3
k <- 3
k3_model <- kmeans(features_scaled, centers = k, nstart = 25)

# Silhouette object
sil_obj <- silhouette(k3_model$cluster, dist(features_scaled))

# Average silhouette width (you can quote this number in your text)
mean_silhouette <- mean(sil_obj[, "sil_width"])
mean_silhouette

# Plot silhouette for K = 3
fviz_silhouette(sil_obj) +
  labs(title = "Silhouette Plot for K = 3")
```

The overall average Silhouette score was approximately 0.97, which
indicates excellent clustering performance. Cluster 2 had a Silhouette
width of 0.98, showing extremely tight grouping and strong separation —
consistent with its large count of low-sales items. Clusters 1 and 3 had
lower Silhouette values (0.19 and 0.09), which is expected for smaller
high-value and mid-value groups that lie further away from the large
slow-moving cluster.

The very high overall Silhouette score confirms that 𝐾=3 forms
well-separated and meaningful clusters.

## Cluster Profiling(Sizes + Centroids + Summary)

Once clustering was completed, the characteristics of each cluster were
examined using the cluster sizes, centroid values (scaled units), and
summary statistics (original units). This profiling step helps translate
numerical cluster assignments into meaningful business categories.

```{r}
# Add cluster labels back to item data
items_clustered <- items %>%
  mutate(cluster = factor(k3_model$cluster))
```

```{r}
# Cluster sizes (how many items in each cluster)
cluster_sizes <- items_clustered %>%
  count(cluster)

cluster_sizes
```
The cluster size results showed a very uneven distribution:

-   Cluster 1: 17 items

-   Cluster 2: 37,667 items

-   Cluster 3: 240 items
```{r}
# Centroids on scaled features (for technical report)
cluster_centroids_scaled <- as_tibble(k3_model$centers) %>%
  mutate(cluster = factor(1:3))

cluster_centroids_scaled
```

```{r}

# Summary stats in original units (mean and median)
cluster_summary <- items_clustered %>%
  group_by(cluster) %>%
  summarise(
    mean_total_sales   = mean(TOTAL_SALES, na.rm = TRUE),
    median_total_sales = median(TOTAL_SALES, na.rm = TRUE),
    mean_retail_sales  = mean(RETAIL_SALES, na.rm = TRUE),
    mean_wh_sales      = mean(WAREHOUSE_SALES, na.rm = TRUE),
    .groups = "drop"
  )

cluster_summary
```



The centroids and averages revealed that:

-   Cluster 1 has the highest total sales (≈139,000 on average), making
    it the Fast-Moving cluster.

-   Cluster 2 has very low total sales (≈155), representing Slow-Moving
    items.

-   Cluster 3 shows moderate total sales (≈16,695), identifying it as
    the Moderate-Moving cluster.

This three-group structure matches typical inventory movement patterns
in retail and aligns with managerial frameworks for stock
classification. The profiling confirms that K-Means successfully
distinguished high-demand items from mid-level and slow-moving products.

## Exploratory Analysis

Before applying K-Means, the dataset was cleaned and summarized to
examine overall sales trends, supplier performance, and product
movement. Using RStudio and packages such as tidyverse, ggplot2, and
lubridate, we computed descriptive statistics and generated exploratory
visualizations.

Example snippet of the cleaned data (aggregated at the item level)
includes variables like:

-   YEAR, MONTH

-   SUPPLIER, ITEM CODE, ITEM DESCRIPTION, ITEM TYPE

-   RETAIL SALES, RETAIL TRANSFERS, WAREHOUSE SALES

-   TOTAL_SALES

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)


# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)

# Read Excel file (update the file path if needed)

data <- read_excel("Retail_Sales.xlsx")







# Remove duplicate rows
data <- distinct(data)

# Handle missing values (replace or remove)
data <- data %>%
  mutate(
    SUPPLIER = ifelse(is.na(SUPPLIER), "Unknown Supplier", SUPPLIER),
    `ITEM TYPE` = ifelse(is.na(`ITEM TYPE`), "Unknown Type", `ITEM TYPE`),
    `RETAIL SALES` = ifelse(is.na(`RETAIL SALES`), 0, `RETAIL SALES`),
    `RETAIL TRANSFERS` = ifelse(is.na(`RETAIL TRANSFERS`), 0, `RETAIL TRANSFERS`),
    `WAREHOUSE SALES` = ifelse(is.na(`WAREHOUSE SALES`), 0, `WAREHOUSE SALES`)
  )

# Create a Total Sales column
data <- data %>%
  mutate(TOTAL_SALES = `RETAIL SALES` + `RETAIL TRANSFERS` + `WAREHOUSE SALES`)

# Combine YEAR and MONTH into a proper Date column
data <- data %>%
  mutate(Date = make_date(YEAR, MONTH, 1))

# View first few cleaned rows in a table
head(data, 20) %>% 
  knitr::kable(caption = "first few Cleaned Retail Sales Data")
```

```{r}
# Total Sales Trend Over Time

library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)

# Ensure the Date column is in date format
data <- data %>%
  mutate(Date = as.Date(Date))

# Aggregate sales by date (if multiple transactions per day)
daily_sales <- data %>%
  group_by(Date) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

# Line chart for total sales trend
ggplot(daily_sales, aes(x = Date, y = Total_Sales)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "orange", size = 1) +
  labs(
    title = "Total Sales Trend Over Time",
    x = "Date of Transaction",
    y = "Total Daily Sales (in USD)"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

**Insights:** This visualization shows the daily total sales trend over
time. It highlights how sales fluctuate from day to day, revealing
patterns such as peaks during high-demand periods and drops during
slower days. The trend line helps identify overall performance changes
and potential seasonal variations in sales.

```{r}
# Top 10 Suppliers by Total Sales

top_suppliers <- data %>%
  group_by(SUPPLIER) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

ggplot(top_suppliers, aes(x = reorder(SUPPLIER, Total_Sales), y = Total_Sales, fill = SUPPLIER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Suppliers by Total Sales", x = "Supplier", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**A few suppliers control the majority of total revenue,
according to the sales analysis. - With the greatest sales, Crown
Imports, Miller Brewing Company, and Anheuser Busch Inc. demonstrate a
concentrated supplier base where major vendors control the majority of
the market. For these major contributors, this pattern emphasizes the
significance of keeping solid supplier relationships and precise demand
forecasting.

```{r}
# Sales by Item Type

ggplot(data, aes(x = `ITEM TYPE`, y = TOTAL_SALES, fill = `ITEM TYPE`)) +
  geom_bar(stat = "summary", fun = "sum") +
  labs(title = "Total Sales by Item Type", x = "Item Type", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**Beer dominates the retail industry, with sales close to \$8
million, greatly outpacing all other categories, according to the sales
distribution across item kinds. Liquor and Wine occupy the second and
third spots, reflecting moderate but sustained demand. The fact that
categories like dunnage, kegs, and non-alcoholic items make up very
little of the overall revenue highlights the fact that alcoholic
beverages are the main driver of the county's retail sector.

-   Beer's dominance suggests a fast-moving inventory sector due to
    frequent use and high brand loyalty. However, goods classified as
    "Kegs" and "Non-Alcoholic" move slowly and would need focused
    marketing techniques to boost sales.

```{r}
# Comparison: Retail vs Warehouse Sales

sales_compare <- data %>%
  summarise(
    Retail_Sales = sum(`RETAIL SALES`, na.rm = TRUE),
    Warehouse_Sales = sum(`WAREHOUSE SALES`, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Category", values_to = "Sales")

ggplot(sales_compare, aes(x = Category, y = Sales, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Retail vs Warehouse Sales", x = "", y = "Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Insights:**Warehouse sales are almost three times greater than retail
sales, according to the analysis. This implies that rather than being
sold directly to customers, the majority of merchandise is transferred
in large quantities via centralized distribution. Although this type of
layout promotes productivity, careful inventory balancing is necessary
to prevent overstocking.

```{r}

# Monthly Sales Heatmap
library(viridis)

monthly_sales <- data %>%
  group_by(YEAR, MONTH) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

ggplot(monthly_sales, aes(x = factor(MONTH), y = factor(YEAR), fill = Total_Sales)) +
  geom_tile() +
  scale_fill_viridis(option = "magma", direction = -1, labels = comma) +
  labs(title = "Monthly Sales Heatmap", x = "Month", y = "Year", fill = "Total Sales") +
  theme_minimal()

```

**Insights:**Due to seasonal and holiday demand spikes, sales activity
peaks between December and the middle of the year. Early months see
lower sales, while anomalies in 2020 point to potential outside
interruptions such as the COVID-19 pandemic. Planning stock levels
around seasonal trends is made easier by identifying these patterns.

# Modeling and Results

## Modeling and Data Preparation

In order to eliminate duplicates, fill in missing values, and normalize
variables for analysis, the Montgomery County retail dataset performed
preprocessing. Retail, warehouse, and transfer sales were combined to
produce a new Total Sales feature that made sure every record
represented all transaction activity. The important quantitative
variables—Total Sales, Retail Sales, and Warehouse Sales—were then
grouped based on their sales volume and consistency using K-Means
clustering.

K = 3 was found to be the ideal number of clusters using the Elbow and
Silhouette methods, creating three separate groups that represented
Fast-Moving, Moderate-Moving, and Slow-Moving objects.

## Cluster Formation and Categorization

After running K-Means with 𝐾=3, the items in the dataset were divided
into three distinct clusters based on their Total Sales, Retail Sales,
and Warehouse Sales values.

Using the cluster summary results:

-   Cluster 1 has only 17 items, but an average total sales of about
    138,991, with very high warehouse sales.

-   Cluster 2 has 37,667 items, but a very low average total sales of
    about 155, and low retail and warehouse sales.

-   Cluster 3 has 240 items, with an average total sales of around
    16,695, and moderate warehouse and retail values.

Based on these patterns, the clusters were interpreted as:

-   Cluster 1 – Fast-Moving Items Items in this cluster have extremely
    high total and warehouse sales. They represent a small set of “core”
    products that sell in large volumes and are critical for revenue.
    These are high-demand items that require frequent replenishment and
    careful monitoring to avoid stock-outs.

-   Cluster 2 – Slow-Moving Items This cluster contains the majority of
    items, but with very low total sales on average. Many of these
    products have limited or irregular demand, and some may be niche,
    specialty, or rarely purchased items. They have low retail and
    warehouse movement compared to the other clusters.

-   Cluster 3 – Moderate-Moving Items Items in this cluster occupy the
    middle ground. Their total sales are much higher than slow movers
    but lower than the fast movers in Cluster 1. They show relatively
    steady and predictable performance over time. These products are
    often wines and mid-tier spirits that generate regular but not
    extreme turnover.

This structure matches the typical inventory classification into fast-,
moderate-, and slow-moving items and confirms that K-Means is capturing
meaningful differences in sales behavior.

## Visualization-Based Evidence

1.**Top Suppliers by Total Sales:**A bar chart of total sales by
supplier showed that companies such as Crown Imports, Miller Brewing
Company, and Anheuser Busch Inc. account for a large share of overall
revenue. Many of their key products fall into Cluster 1 (Fast-Moving),
which is consistent with the very high total sales and strong warehouse
movement observed in that cluster. In contrast, smaller suppliers tend
to have more products in Cluster 2 (Slow-Moving) or Cluster 3
(Moderate-Moving).

2.**Total Sales by Item Type:**The visualization of total sales by item
type showed that beer dominates total sales, followed by liquor and
wine. Fast-moving items in Cluster 1 are heavily concentrated in
high-volume categories like beer and popular liquor brands. Cluster 3
(Moderate-Moving) contains many wine and mid-tier products with steady
but lower sales. Categories such as kegs, dunnage, and non-alcoholic
items are more frequently found in Cluster 2 (Slow-Moving), reflecting
their low or irregular turnover.

3.**Retail vs Warehouse Sales:** When comparing total retail and
warehouse sales, we see that warehouse sales (\~\$7.8M) are much larger
than retail sales (\~\$2.1M). Items in Cluster 1 (Fast-Moving) tend to
have high warehouse sales and noticeable retail movement, indicating
that these products are stocked in bulk to meet continuous demand. Items
in Cluster 3 (Moderate-Moving) show mid-level warehouse and retail
sales. Many items in Cluster 2 (Slow-Moving) have low values in both
channels and appear only occasionally in sales records.

4.**Heatmap of Monthly Sales:** A monthly heatmap of sales activity
shows that fast-moving items (Cluster 1) have strong seasonal peaks,
especially in summer and December, but also maintain significant
baseline sales throughout the year. Moderate-moving items (Cluster 3)
show smoother, stable patterns across months. Slow-moving items (Cluster
2) exhibit many months with very low or zero sales, with only occasional
spikes. The irregular behavior in 2020 suggests potential external
influences such as COVID-19.

Together, these visual patterns align well with the numeric cluster
profiles and strengthen the interpretation of Clusters 1, 2, and 3 as
fast, slow, and moderate moving groups.

## Cluster Insights

**Fast-Moving Items (Cluster 1)** Fast-moving items are a small group of
products with extremely high total and warehouse sales. These include
major beer brands and top liquor products supplied by key vendors like
Crown Imports and Miller Brewing. Because they generate a large portion
of total revenue, they must be managed carefully:

-   They require frequent replenishment to prevent stock-outs.

-   Forecasting errors can have a big impact on customer satisfaction
    and sales.

-   These items should receive priority in storage, shelf space, and
    monitoring.

**Slow-Moving Items (Cluster 2)** Slow-moving items form the largest
cluster, but each item sells in small quantities. Many products in this
group are niche, seasonal, or low-demand items, including some kegs and
non-alcoholic products:

-   They are at higher risk of overstocking and inventory holding costs.

-   Managers may consider reducing order quantities, reviewing
    assortment, or using promotions and discounts to move slow stock.

-   Long-term slow movers may be candidates for discontinuation if they
    do not support strategic goals.

**Moderate-Moving Items (Cluster 3)** Moderate-moving items sit between
the other two clusters. They have mid-range total sales and stable,
predictable demand. Many wines and mid-tier spirits fall into this
category:

-   They provide steady revenue and are important for assortment
    breadth.

-   They benefit from balanced safety stock—not as aggressive as fast
    movers, but also not as restricted as slow movers.

-   Regular performance reviews can help decide whether some items
    should be promoted into a more active category or reduced if demand
    declines.

Overall, the three clusters give a clear structure to the product
portfolio. Managers can focus their attention where it matters most:
protecting availability for fast-moving items, keeping moderate items
stable, and controlling costs for slow-moving items.

# Conclusion

## Key Findings

-   The K-Means clustering divided the products into three clear groups
    — Fast-Moving, Moderate-Moving, and Slow-Moving items.

-   Fast-Moving items, like beer and popular liquor brands from
    suppliers such as Crown Imports and Miller Brewing, had very high
    and steady sales. These products need to be restocked often.

-   Moderate-Moving items, such as wine, showed average and regular
    sales. They are stable products that need balanced inventory levels.

-   Slow-Moving items had low or irregular sales, mostly niche or
    seasonal goods. These products need careful monitoring to avoid
    overstock and storage costs.

-   Sales increased during holiday seasons like summer and December,
    showing that demand changes with time and events.

## Results Summary

The clustering results helped group inventory based on sales behavior
and stock movement.

This makes it easier for managers to:

-   Focus on fast-selling items to keep shelves full,

-   Maintain enough stock of moderate-selling items, and

-   Reduce extra inventory of slow-selling items.

Using K-Means clustering for inventory helps stores plan better, avoid
waste, and respond quickly to customer demand. Overall, it supports
smarter inventory control, better forecasting, and more efficient retail
operations.
