---
title: "Present a great story for data science projects"
subtitle: "This is a Slide Template"
author: "Students names (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction

Retail businesses generate massive amounts of sales and customer data.

To make sense of this data, companies increasingly use clustering
methods like K-Means to identify hidden patterns in purchasing behavior
and product movement.

Research shows that K-Means helps improve:

-   Customer segmentation

-   Inventory forecasting

-   Marketing targeting

-   Demand planning

Studies across retail markets—from large chains to small grocery
stores—show that K-Means can separate products into meaningful groups
such as fast, moderate, and slow movers.

## Introduction

Improvements like weighted K-Means and MiniBatch K-Means have made
clustering faster and more accurate, especially for large datasets.

However, many past studies do not clearly report preprocessing steps or
cluster validation.

This project fills those gaps by:

-   Using a transparent, publicly available government dataset

-   Carefully cleaning and preparing the data

-   Choosing the best number of clusters using both statistics and
    business logic

-   Validating the results with Silhouette and CH Index

## Problem & Research Questions

**Problem**

-   Overstocking slow movers → wasted space and cost

-   Understocking fast movers → stock-outs and lost sales

**Research Questions**

Can K-Means find meaningful clusters in the county’s alcohol sales?

Do clusters map to:

-   Fast-moving

-   Moderate-moving

-   Slow-moving items?

How can these clusters help managers make better decisions?

## Literature Review

K-Means is widely used in retail analytics:

-   Customer segmentation

-   Product grouping

-   Demand forecasting

@prasetiya2025review:

-   K-Means + association rules helps targeted marketing.

-   Works well on large retail datasets.

@omol2024kenya:

-   K-Means for grocery stores in Kenya.

-   Customer behavior linked to age and income.

@bala2012forecast:

-   Clustering + demand forecasting → better inventory accuracy.

@sitorus2023inventory:

-   Products clustered into fast, medium, slow demand categories.

## Literature Review

**Gaps on Many papers:**

-   Do not clearly show data preprocessing.

-   Do not fully explain how K is chosen.

-   Use non-public datasets.

**Our contribution**

Use an open government dataset.

Clearly document:

-   Cleaning and feature engineering

-   Choosing K

-   Cluster validation and interpretation

## Methods: K-Means Overview

K-Means is an unsupervised machine learning algorithm that groups
similar data points into K clusters. Each cluster has a centroid, and
items are assigned to the cluster whose centroid they are closest to.

Mathematically, K-Means works by minimizing the Within-Cluster Sum of
Squares (WCSS):

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$

The goal is to create clusters that are compact (low variance) and well
separated.

In this project, K-Means is used to group retail items based on their
sales behavior—helping identify fast-, moderate-, and slow-moving
products.

Why we use it:

-   Finds natural patterns in sales data

-   Helps classify product movement levels

-   Supports better inventory planning

-   Simplifies complex data into clear groups

**Simple Explanation:** K-Means automatically divides items into groups
based on how they sell. This helps managers quickly understand which
products sell fast, which are steady, and which move slowly.

## Methods: Choosing K

We used three approaches:

**Elbow Method** [@MacQueen1967]

-   Plot WCSS vs. K

-   Look for the “elbow” where improvement slows down $$
    WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
    $$

**Silhouette Coefficient** [@Rousseeuw1987]

Measures how well items fit into their assigned cluster vs neighbors

Range: -1 to 1 (higher is better) $$
s(i) = \frac{b(i) - a(i)}{\max(a(i), b(i))}
$$

## Methods: Choosing K

**Calinski–Harabasz Index** Measures how well clusters are formed by
comparing:

-   Between-cluster variance (how far clusters are from each other)

-   Within-cluster variance (how compact each cluster is)

Higher CH values = better, clearer clusters.

**Why It Works for Our Data**

CH scores peaked at K = 3, showing strong separation and tight grouping.

This also fits business logic of Fast, Moderate, and Slow-moving items.

**Takeaway:** CH Index confirms that 3 clusters give the most meaningful
and well-defined structure.

## Data Source

Dataset: Warehouse and Retail Sales

-   Montgomery County Open Data Portal
    [@MontgomeryCountyWarehouseRetail2025]

Period: 2017–2020

Records include:

-   Supplier name

-   Item code and item type (beer, wine, liquor, etc.)

-   Retail sales, retail transfers, warehouse sales

-   Year and month

Public-domain dataset → analysis is reproducible

## Data Preprocessing

Aggregated data from monthly level to item level:

-   One row per supplier–item.

Created:

-   TOTAL_SALES = RETAIL_SALES + RETAIL_TRANSFERS + WAREHOUSE_SALES

Removed:

-   Duplicates

-   Items with TOTAL_SALES = 0

Handled missing values:

-   Replaced NAs in sales with 0

Standardized numeric features for clustering:

-   TOTAL_SALES, RETAIL_SALES, WAREHOUSE_SALES

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(cluster)    # for silhouette()
library(factoextra) # for convenient plots (optional)
library(scales)   

# Read Excel file
retail_raw <- read_excel("Retail_Sales.xlsx")

# Aggregate to item level: one row per supplier-item
items <- retail_raw %>%
  group_by(SUPPLIER, `ITEM CODE`, `ITEM DESCRIPTION`, `ITEM TYPE`) %>%
  summarise(
    RETAIL_SALES     = sum(`RETAIL SALES`, na.rm = TRUE),
    RETAIL_TRANSFERS = sum(`RETAIL TRANSFERS`, na.rm = TRUE),
    WAREHOUSE_SALES  = sum(`WAREHOUSE SALES`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    TOTAL_SALES = RETAIL_SALES + RETAIL_TRANSFERS + WAREHOUSE_SALES
  ) %>%
  filter(TOTAL_SALES > 0)  # keep only items that actually sold something

# Numeric features for clustering
features <- items %>%
  select(TOTAL_SALES, RETAIL_SALES, WAREHOUSE_SALES)

# Scale features
features_scaled <- scale(features)
```

## Elbow Method

The Elbow Method was used to evaluate different values of 𝐾 by
calculating the Within-Cluster Sum of Squares (WCSS) for 𝐾=2 through
𝐾=8. The Elbow plot shows how much WCSS decreases as the number of
clusters increases. A sharp drop is usually followed by a flattening of
the curve—this “elbow” indicates a good balance between simplicity and
accuracy.

```{r}
set.seed(123)

k_values <- 2:8

wcss <- map_dbl(k_values, function(k) {
  kmeans(features_scaled, centers = k, nstart = 10)$tot.withinss
})

elbow_df <- tibble(
  k = k_values,
  WCSS = wcss
)

ggplot(elbow_df, aes(x = k, y = WCSS)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Elbow Plot for Choosing K",
    x = "Number of Clusters (K)",
    y = "Total Within-Cluster Sum of Squares (WCSS)"
  ) +
  theme_minimal()
```

The WCSS curve showed a clear bend at 𝐾=3. Beyond three clusters, the
reduction in WCSS became minimal, suggesting that additional clusters
would not significantly improve separation. This result supports
choosing three clusters as an appropriate and efficient segmentation of
the retail items.

## Silhouette Analysis

To validate the Elbow Method, a Silhouette analysis was performed for
𝐾=3. The Silhouette score measures how close each item is to its own
cluster compared to other clusters. Values close to +1 indicate strong
cohesion within clusters and clear separation between them.

```{r}
set.seed(123)

set.seed(123)

# Fit K-Means with K = 3
k <- 3
k3_model <- kmeans(features_scaled, centers = k, nstart = 25)

# Silhouette object
sil_obj <- silhouette(k3_model$cluster, dist(features_scaled))

# Average silhouette width (you can quote this number in your text)
mean_silhouette <- mean(sil_obj[, "sil_width"])
mean_silhouette
# Plot silhouette for K = 3
fviz_silhouette(sil_obj) +
  labs(title = "Silhouette Plot for K = 3")
```

-   Average Silhouette ≈ 0.97 → excellent cluster structure.

-   Cluster 2: Silhouette ≈ 0.98 → very tight and well-separated.

-   Clusters 1 and 3: smaller groups with lower but positive
    silhouettes.

-   Confirms that 3 clusters are meaningful and well-separated.

## Cluster Profiling: Sizes

```{r}
items_clustered <- items %>%
mutate(cluster = factor(k3_model$cluster))

# Cluster sizes (how many items in each cluster)
cluster_sizes <- items_clustered %>%
  count(cluster)

cluster_sizes

```

Cluster sizes

-   Cluster 1: 17 items

-   Cluster 2: 37,667 items

-   Cluster 3: 240 items

This suggests:

-   A small set of very high-volume items (Cluster 1),

-   A large pool of low-volume items (Cluster 2),

-   A mid-sized group with moderate sales (Cluster 3).

## Cluster Profiling: Means

```{r}
# Centroids on scaled features (for technical report)
cluster_centroids_scaled <- as_tibble(k3_model$centers) %>%
  mutate(cluster = factor(1:3))

cluster_centroids_scaled
```

```{r}
# Summary stats in original units (mean and median)
cluster_summary <- items_clustered %>%
  group_by(cluster) %>%
  summarise(
    mean_total_sales   = mean(TOTAL_SALES, na.rm = TRUE),
    median_total_sales = median(TOTAL_SALES, na.rm = TRUE),
    mean_retail_sales  = mean(RETAIL_SALES, na.rm = TRUE),
    mean_wh_sales      = mean(WAREHOUSE_SALES, na.rm = TRUE),
    .groups = "drop"
  )

cluster_summary
```

**Interpretation** Cluster 1

-   Very high mean total & warehouse sales → Fast-Moving

Cluster 2

-   Very low mean total sales → Slow-Moving

Cluster 3

-   Medium mean total sales → Moderate-Moving

## Exploratory Analysis: Total Sales Trend

```{r}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)


# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)

# Read Excel file (update the file path if needed)

data <- read_excel("Retail_Sales.xlsx")







# Remove duplicate rows
data <- distinct(data)

# Handle missing values (replace or remove)
data <- data %>%
  mutate(
    SUPPLIER = ifelse(is.na(SUPPLIER), "Unknown Supplier", SUPPLIER),
    `ITEM TYPE` = ifelse(is.na(`ITEM TYPE`), "Unknown Type", `ITEM TYPE`),
    `RETAIL SALES` = ifelse(is.na(`RETAIL SALES`), 0, `RETAIL SALES`),
    `RETAIL TRANSFERS` = ifelse(is.na(`RETAIL TRANSFERS`), 0, `RETAIL TRANSFERS`),
    `WAREHOUSE SALES` = ifelse(is.na(`WAREHOUSE SALES`), 0, `WAREHOUSE SALES`)
  )

# Create a Total Sales column
data <- data %>%
  mutate(TOTAL_SALES = `RETAIL SALES` + `RETAIL TRANSFERS` + `WAREHOUSE SALES`)

# Combine YEAR and MONTH into a proper Date column
data <- data %>%
  mutate(Date = make_date(YEAR, MONTH, 1))


# Total Sales Trend Over Time

library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)

# Ensure the Date column is in date format
data <- data %>%
  mutate(Date = as.Date(Date))

# Aggregate sales by date (if multiple transactions per day)
daily_sales <- data %>%
  group_by(Date) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

# Line chart for total sales trend
ggplot(daily_sales, aes(x = Date, y = Total_Sales)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "orange", size = 1) +
  labs(
    title = "Total Sales Trend Over Time",
    x = "Date of Transaction",
    y = "Total Daily Sales (in USD)"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

  
```

**Key takeaway**

Clear seasonal patterns:

-   Peaks around mid-year and December.

-   Dips in early months.

2020 shows irregularities → likely COVID-19 effect.

## Exploratory Analysis: Top Suppliers

```{r}
top_suppliers <- data %>%
  group_by(SUPPLIER) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

ggplot(top_suppliers, aes(x = reorder(SUPPLIER, Total_Sales), y = Total_Sales, fill = SUPPLIER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Suppliers by Total Sales", x = "Supplier", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")
```
**Key takeaway**
-   A few suppliers (e.g., Crown Imports, Miller, Anheuser Busch)
    dominate sales.

-   These suppliers contribute heavily to fast-moving items.

## Exploratory Analysis: Item Types

```{r}
# Sales by Item Type

ggplot(data, aes(x = `ITEM TYPE`, y = TOTAL_SALES, fill = `ITEM TYPE`)) +
  geom_bar(stat = "summary", fun = "sum") +
  labs(title = "Total Sales by Item Type", x = "Item Type", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Key takeaway**

- Beer dominates total sales.

- Liquor and wine are moderate.

- Kegs and non-alcoholic products → slow-moving side.

## Cluster Visualization: Scatter Plot
```{r}
ggplot(items_clustered,
       aes(x = WAREHOUSE_SALES,
           y = TOTAL_SALES,
           color = cluster)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "K-Means Clusters (K = 3)",
    x = "Warehouse Sales",
    y = "Total Sales",
    color = "Cluster"
  ) +
  theme_minimal()

```
**Interpretation**

- Cluster 1: points in high-sales region → Fast-Moving.

- Cluster 2: dense cloud near origin → Slow-Moving.

- Cluster 3: middle zone → Moderate-Moving.


## Cluster Insights: Fast-Moving (Cluster 1)


Very high total and warehouse sales.

Small number of items, but critical for revenue.

Mostly:

- Major beer brands

- Top liquor products

Management actions:

- Frequent replenishment

- Tight forecasting

- Priority storage & shelf space

## Cluster Insights: Slow-Moving (Cluster 2)
Largest number of items, but low sales per item.

Many are:

- Niche

- Seasonal

- Low-demand products

Risks:

- Overstocking

- High holding costs

Management actions:

- Reduce order quantities

- Use promotions/discounts

- Consider discontinuation of chronic slow movers

## Cluster Insights: Moderate-Moving (Cluster 3)
Medium total sales, stable demand.

Many wines and mid-tier spirits.

Role:

- Provide assortment breadth

- Steady revenue stream

Management actions:

- Balanced safety stock

Regular review of trends

- Adjust assortment based on performance

## Conclusion: Key Findings

K-Means with K = 3 produced three meaningful clusters:

- Cluster 1: Fast-moving items

- Cluster 2: Slow-moving items

- Cluster 3: Moderate-moving items

Seasonal patterns and supplier concentration strongly affect sales.

Clustering turned a large, complex dataset into three clear groups
that managers can use for decision-making.

## Conclusion: Managerial Implications
Focus on:

- Fast-moving items → avoid stock-outs.

- Moderate-moving items → maintain stable levels.

- Slow-moving items → control inventory and reduce waste.

Using K-Means for inventory:

- Improves planning and forecasting

- Supports data-driven retail strategy

- Reduces risk of overstock and lost sales

## References
