---
title: "Applying K-Means Clustering in retail industry"

author: "Charan Gullipalli, Lavanya Guntupalli, Arun Pandian  (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
      
   
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---
<style>
.reveal {
  background-color: #ADD8E6 !important;  /* sky blue */
}

.reveal .slide-background {
  background-color: #ADD8E6 !important;  /* sky blue */
}
</style>
## Introduction {.smaller}

Retail businesses generate massive amounts of sales and customer data.

To make sense of this data, companies increasingly use clustering
methods like K-Means to identify hidden patterns in purchasing behavior
and product movement.

Research shows that K-Means helps improve:

-   Customer segmentation

-   Inventory forecasting

-   Marketing targeting

-   Demand planning

Studies across retail markets—from large chains to small grocery
stores—show that K-Means can separate products into meaningful groups
such as fast, moderate, and slow movers.

## Introduction {.smaller}

Improvements like weighted K-Means and MiniBatch K-Means have made
clustering faster and more accurate, especially for large datasets.

However, many past studies do not clearly report preprocessing steps or
cluster validation.

This project fills those gaps by:

-   Using a transparent, publicly available government dataset

-   Carefully cleaning and preparing the data

-   Choosing the best number of clusters using both statistics and
    business logic

-   Validating the results with Silhouette and CH Index

## Problem & Research Questions {.smaller}

**Problem**

-   Overstocking slow movers → wasted space and cost

-   Understocking fast movers → stock-outs and lost sales

**Research Questions**

Can K-Means find meaningful clusters in the county’s alcohol sales?

Do clusters map to:

-   Fast-moving

-   Moderate-moving

-   Slow-moving items?

How can these clusters help managers make better decisions?

## Literature Review {.smaller}

K-Means is widely used in retail analytics:

-   Customer segmentation

-   Product grouping

-   Demand forecasting

@prasetiya2025review:

-   K-Means + association rules helps targeted marketing.

-   Works well on large retail datasets.

@omol2024kenya:

-   K-Means for grocery stores in Kenya.

-   Customer behavior linked to age and income.

## Literature Review {.smaller}

@bala2012forecast:

-   Clustering + demand forecasting → better inventory accuracy.

@sitorus2023inventory:

-   Products clustered into fast, medium, slow demand categories.

## Literature Review {.smaller}

**Gaps on Many papers:**
-   Do not clearly show data preprocessing.

-   Do not fully explain how K is chosen.

-   Use non-public datasets.

**Our contribution**
Use an open government dataset.

Clearly document:

-   Cleaning and feature engineering

-   Choosing K

-   Cluster validation and interpretation

## Methods: K-Means Overview {.smaller}

K-Means is an unsupervised machine learning algorithm that groups
similar data points into K clusters. Each cluster has a centroid, and
items are assigned to the cluster whose centroid they are closest to.

Mathematically, K-Means works by minimizing the Within-Cluster Sum of
Squares (WCSS):

$$
WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
$$

The goal is to create clusters that are compact (low variance) and well
separated.

In this project, K-Means is used to group retail items based on their
sales behavior—helping identify fast-, moderate-, and slow-moving
products.

Why we use it:
-   Finds natural patterns in sales data

-   Helps classify product movement levels

-   Supports better inventory planning

-   Simplifies complex data into clear groups

**Simple Explanation:** K-Means automatically divides items into groups
based on how they sell. This helps managers quickly understand which
products sell fast, which are steady, and which move slowly.

## Methods: Choosing K {.smaller}

We used three approaches:

**Elbow Method** [@MacQueen1967]

-   Plot WCSS vs. K

-   Look for the “elbow” where improvement slows down $$
    WCSS = \sum_{k=1}^{K} \sum_{x_i \in C_k} \| x_i - \mu_k \|^2
    $$

**Silhouette Coefficient** [@Rousseeuw1987]

Measures how well items fit into their assigned cluster vs neighbors

Range: -1 to 1 (higher is better) $$
s(i) = \frac{b(i) - a(i)}{\max(a(i), b(i))}
$$

## Methods: Choosing K {.smaller}

**Calinski–Harabasz Index** Measures how well clusters are formed by
comparing:(@Calinski1974)

-   Between-cluster variance (how far clusters are from each other)

-   Within-cluster variance (how compact each cluster is)

Higher CH values = better, clearer clusters.

**Why It Works for Our Data**

CH scores peaked at K = 3, showing strong separation and tight grouping.

This also fits business logic of Fast, Moderate, and Slow-moving items.

**Takeaway:** CH Index confirms that 3 clusters give the most meaningful
and well-defined structure.

## Data Source {.smaller}

Dataset: Warehouse and Retail Sales

-   Montgomery County Open Data Portal
    [@MontgomeryCountyWarehouseRetail2025]

Period: 2017–2020

Records include:

-   Supplier name

-   Item code and item type (beer, wine, liquor, etc.)

-   Retail sales, retail transfers, warehouse sales

-   Year and month

Public-domain dataset → analysis is reproducible

## Data Preprocessing {.smaller}

Aggregated data from monthly level to item level:
-   One row per supplier–item.

Created:

-   TOTAL_SALES = RETAIL_SALES + RETAIL_TRANSFERS + WAREHOUSE_SALES

Removed:

-   Duplicates

-   Items with TOTAL_SALES = 0

Handled missing values:

-   Replaced NAs in sales with 0

Standardized numeric features for clustering:

-   TOTAL_SALES, RETAIL_SALES, WAREHOUSE_SALES

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(cluster)    # for silhouette()
library(factoextra) # for convenient plots (optional)
library(scales)   

# Read Excel file
retail_raw <- read_excel("Retail_Sales.xlsx")

# Aggregate to item level: one row per supplier-item
items <- retail_raw %>%
  group_by(SUPPLIER, `ITEM CODE`, `ITEM DESCRIPTION`, `ITEM TYPE`) %>%
  summarise(
    RETAIL_SALES     = sum(`RETAIL SALES`, na.rm = TRUE),
    RETAIL_TRANSFERS = sum(`RETAIL TRANSFERS`, na.rm = TRUE),
    WAREHOUSE_SALES  = sum(`WAREHOUSE SALES`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    TOTAL_SALES = RETAIL_SALES + RETAIL_TRANSFERS + WAREHOUSE_SALES
  ) %>%
  filter(TOTAL_SALES > 0)  # keep only items that actually sold something

# Numeric features for clustering
features <- items %>%
  select(TOTAL_SALES, RETAIL_SALES, WAREHOUSE_SALES)

# Scale features
features_scaled <- scale(features)
```

## Elbow Method {.smaller}
The Elbow Method evaluates different values of K by computing the Within-Cluster Sum of Squares (WCSS) for each K. As K increases, WCSS decreases because clusters become smaller. The “elbow point” is where the reduction in WCSS slows down, indicating the best balance between simplicity and accuracy.

```{r}
set.seed(123)

k_values <- 2:8

wcss <- map_dbl(k_values, function(k) {
  kmeans(features_scaled, centers = k, nstart = 10)$tot.withinss
})

elbow_df <- tibble(
  k = k_values,
  WCSS = wcss
)

ggplot(elbow_df, aes(x = k, y = WCSS)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Elbow Plot for Choosing K",
    x = "Number of Clusters (K)",
    y = "Total Within-Cluster Sum of Squares (WCSS)"
  ) +
  theme_minimal()
```

In our plot, the curve shows a clear bend at K = 3, meaning that adding more clusters does not significantly reduce WCSS. Therefore, K = 3 was selected as the optimal and most efficient number of clusters for this dataset.

## Silhouette Analysis {.smaller}
A Silhouette analysis was used to validate K = 3 by measuring how well each item fits within its assigned cluster compared to others. Scores close to +1 indicate strong cohesion and clear separation. The high average silhouette value confirms that K = 3 forms well-defined, meaningful clusters.
```{r}
set.seed(123)

set.seed(123)

# Fit K-Means with K = 3
k <- 3
k3_model <- kmeans(features_scaled, centers = k, nstart = 25)

# Silhouette object
sil_obj <- silhouette(k3_model$cluster, dist(features_scaled))

# Average silhouette width (you can quote this number in your text)
mean_silhouette <- mean(sil_obj[, "sil_width"])
mean_silhouette
# Plot silhouette for K = 3
fviz_silhouette(sil_obj) +
  labs(title = "Silhouette Plot for K = 3")
```
## Silhouette Analysis {.smaller}
-   Average Silhouette ≈ 0.97 → excellent cluster structure.
-   Cluster 2: Silhouette ≈ 0.98 → very tight and well-separated.
-   Clusters 1 and 3: smaller groups with lower but positive
    silhouettes.
-   Confirms that 3 clusters are meaningful and well-separated.



## Cluster Profiling: Sizes {.smaller}

```{r}
items_clustered <- items %>%
mutate(cluster = factor(k3_model$cluster))

# Cluster sizes (how many items in each cluster)
cluster_sizes <- items_clustered %>%
  count(cluster)

cluster_sizes

```

Cluster sizes

-   Cluster 1: 17 items

-   Cluster 2: 37,667 items

-   Cluster 3: 240 items

This suggests:

-   A small set of very high-volume items (Cluster 1),

-   A large pool of low-volume items (Cluster 2),

-   A mid-sized group with moderate sales (Cluster 3).

## Cluster Profiling: Means {.smaller}

```{r}
# Centroids on scaled features (for technical report)
cluster_centroids_scaled <- as_tibble(k3_model$centers) %>%
  mutate(cluster = factor(1:3))

cluster_centroids_scaled
```

```{r}
# Summary stats in original units (mean and median)
cluster_summary <- items_clustered %>%
  group_by(cluster) %>%
  summarise(
    mean_total_sales   = mean(TOTAL_SALES, na.rm = TRUE),
    median_total_sales = median(TOTAL_SALES, na.rm = TRUE),
    mean_retail_sales  = mean(RETAIL_SALES, na.rm = TRUE),
    mean_wh_sales      = mean(WAREHOUSE_SALES, na.rm = TRUE),
    .groups = "drop"
  )

cluster_summary
```

**Interpretation**

- Cluster 1 has the highest total sales (≈139,000 on average), making it the Fast-Moving cluster.

- Cluster 2 has very low total sales (≈155), representing Slow-Moving items.

- Cluster 3 shows moderate total sales (≈16,695), identifying it as the Moderate-Moving cluster.

## Cluster Visualization: Scatter Plot {.smaller}
```{r}
ggplot(items_clustered,
       aes(x = WAREHOUSE_SALES,
           y = TOTAL_SALES,
           color = cluster)) +
  geom_point(alpha = 0.5, size = 1.1) +
  
  # X-axis: log10, start at 100, no gap
  scale_x_log10(
    labels = scales::comma,
    limits = c(100, NA),
    expand = c(0, 0)   # <-- removes the gap at the axis
  ) +
  
  # Y-axis: log10, start at 100, no gap
  scale_y_log10(
    labels = scales::comma,
    limits = c(100, NA),
    expand = c(0, 0)   # <-- removes the gap at the axis
  ) +
  
  labs(
    title = "K-Means Clusters",
    x = "Warehouse Sales",
    y = "Total Sales",
    color = "Cluster"
  ) +
  theme_minimal()

```
**Interpretation**

- Cluster 1: points in high-sales region → Fast-Moving.

- Cluster 2: dense cloud near origin → Slow-Moving.

- Cluster 3: middle zone → Moderate-Moving.

## Exploratory Analysis:Sales Trend {.smaller}

```{r}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)


# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)

# Read Excel file (update the file path if needed)

data <- read_excel("Retail_Sales.xlsx")







# Remove duplicate rows
data <- distinct(data)

# Handle missing values (replace or remove)
data <- data %>%
  mutate(
    SUPPLIER = ifelse(is.na(SUPPLIER), "Unknown Supplier", SUPPLIER),
    `ITEM TYPE` = ifelse(is.na(`ITEM TYPE`), "Unknown Type", `ITEM TYPE`),
    `RETAIL SALES` = ifelse(is.na(`RETAIL SALES`), 0, `RETAIL SALES`),
    `RETAIL TRANSFERS` = ifelse(is.na(`RETAIL TRANSFERS`), 0, `RETAIL TRANSFERS`),
    `WAREHOUSE SALES` = ifelse(is.na(`WAREHOUSE SALES`), 0, `WAREHOUSE SALES`)
  )

# Create a Total Sales column
data <- data %>%
  mutate(TOTAL_SALES = `RETAIL SALES` + `RETAIL TRANSFERS` + `WAREHOUSE SALES`)

# Combine YEAR and MONTH into a proper Date column
data <- data %>%
  mutate(Date = make_date(YEAR, MONTH, 1))


# Total Sales Trend Over Time

library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)

# Ensure the Date column is in date format
data <- data %>%
  mutate(Date = as.Date(Date))

# Aggregate sales by date (if multiple transactions per day)
daily_sales <- data %>%
  group_by(Date) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

# Line chart for total sales trend
ggplot(daily_sales, aes(x = Date, y = Total_Sales)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "orange", size = 1) +
  labs(
    title = "Total Sales Trend Over Time",
    x = "Date of Transaction",
    y = "Total Daily Sales (in USD)"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

  
```

**Key takeaway**
Clear seasonal patterns:

-   Peaks around mid-year and December.

-   Dips in early months.

2020 shows irregularities → likely COVID-19 effect.

## Exploratory Analysis: Top Suppliers {.smaller}

```{r}
top_suppliers <- data %>%
  group_by(SUPPLIER) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE)) %>%
  arrange(desc(Total_Sales)) %>%
  head(10)

ggplot(top_suppliers, aes(x = reorder(SUPPLIER, Total_Sales), y = Total_Sales, fill = SUPPLIER)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Suppliers by Total Sales", x = "Supplier", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")
```
**Key takeaway**

-   A few suppliers (e.g., Crown Imports, Miller, Anheuser Busch)
    dominate sales.

-   These suppliers contribute heavily to fast-moving items.

## Exploratory Analysis: Item Types {.smaller}

```{r}
# Sales by Item Type

ggplot(data, aes(x = `ITEM TYPE`, y = TOTAL_SALES, fill = `ITEM TYPE`)) +
  geom_bar(stat = "summary", fun = "sum") +
  labs(title = "Total Sales by Item Type", x = "Item Type", y = "Total Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Key takeaway**

- Beer dominates total sales.

- Liquor and wine are moderate.

- Kegs and non-alcoholic products → slow-moving side.

## Exploratory Analysis:Retail vs Warehouse Sales {.smaller}

```{r}
# Comparison: Retail vs Warehouse Sales

sales_compare <- data %>%
  summarise(
    Retail_Sales = sum(`RETAIL SALES`, na.rm = TRUE),
    Warehouse_Sales = sum(`WAREHOUSE SALES`, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Category", values_to = "Sales")

ggplot(sales_compare, aes(x = Category, y = Sales, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Retail vs Warehouse Sales", x = "", y = "Sales") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Key takeaway**

- Warehouse sales are nearly three times higher than retail sales, according to the data.  Merchandise is typically distributed in bulk through centralized channels rather than being sold directly to clients.  While this structure increases productivity, it requires careful inventory balance to avoid overstocking.

## Exploratory Analysis:Monthly Sales Heatmap {.smaller}

```{r}
# Monthly Sales Heatmap
library(viridis)

monthly_sales <- data %>%
  group_by(YEAR, MONTH) %>%
  summarise(Total_Sales = sum(TOTAL_SALES, na.rm = TRUE))

ggplot(monthly_sales, aes(x = factor(MONTH), y = factor(YEAR), fill = Total_Sales)) +
  geom_tile() +
  scale_fill_viridis(option = "magma", direction = -1, labels = comma) +
  labs(title = "Monthly Sales Heatmap", x = "Month", y = "Year", fill = "Total Sales") +
  theme_minimal()
```
**Key takeaway**

- Sales peak in summer and December due to seasonal and holiday demand, while early months show lower activity. Unusual patterns in 2020 suggest external disruptions like COVID-19. Recognizing these seasonal shifts helps optimize inventory planning.


## Cluster Insights:Fast-Moving(Cluster 1) {.smaller}
Items in this cluster represent the highest-performing products in the entire dataset.
Although Cluster 1 contains only a small number of items, these products generate exceptionally high total and warehouse sales, making them the core revenue drivers for the county’s retail system.

These are predominantly major beer brands and leading liquor products, which sell consistently in high volumes across months and seasons. Because of their strong and stable demand, they require close operational attention.

Management actions:

- Ensure frequent replenishment to prevent stock-outs, especially during peak seasons.

- Apply tight demand forecasting, as small forecasting errors can create major revenue losses.

- Allocate priority shelf space and warehouse storage, since these items sustain customer satisfaction and overall sales performance.

## Cluster Insights: Slow-Moving (Cluster 2) {.smaller}

Cluster 2 contains the largest number of items, but each item contributes very little to total sales. These products typically have low, irregular, or seasonal demand, and many belong to niche categories such as specialty beverages, certain kegs, or non-alcoholic items.

Because they occupy significant space but contribute minimally to revenue, these items pose inventory risks, including excess holding costs and potential spoilage or obsolescence.

## Cluster Insights: Slow-Moving (Cluster 2) {.smaller}

**Risks**:

- Overstocking, due to slow turnover

- Increased carrying and storage costs

- Shelf space tied up by low-performing items

**Management actions**:

- Reduce order quantities to avoid unnecessary stock accumulation.

- Use discounts, promotions, or bundling to stimulate movement.

- Consider discontinuing chronic slow movers that repeatedly underperform and do not serve strategic purposes.

## Cluster Insights: Moderate-Moving (Cluster 3) {.smaller}

Items in this cluster fall between fast and slow movers. They exhibit moderate total sales with stable, predictable demand patterns. Many products in this group include wines and mid-tier spirits that customers purchase regularly, though in lower volumes than the top-selling items.

Moderate movers help maintain a balanced product assortment. They diversify revenue streams, support customer preferences, and contribute to consistent monthly sales.

## Cluster Insights: Moderate-Moving (Cluster 3) {.smaller}

**Role & characteristics:**

- Provide assortment breadth to accommodate varied customer tastes

- Generate steady, reliable revenue

- Show consistent demand with fewer extreme peaks or drops

**Management actions:**

- Maintain balanced safety stock—not excessive, but sufficient to avoid shortages

- Conduct regular performance reviews to detect upward or downward trends

- Adjust assortment and marketing strategies based on performance signals

## Conclusion: Key Findings {.smaller}

K-Means clustering with K = 3 successfully segmented retail items into Fast-Moving, Moderate-Moving, and Slow-Moving groups.
The analysis showed strong alignment between statistical patterns and real-world retail behavior:

- Cluster 1 captured the high-demand, high-revenue items that drive operational performance

- Cluster 2 contained low-selling items that require cost control and strategic review

- Cluster 3 represented stable, moderate performers that support consistent sales

Seasonal peaks (especially summer and December) and supplier concentration patterns further validated the cluster interpretations.
Overall, clustering simplified a large and complex dataset into three clear categories that directly support business decision-making.

## Conclusion: Managerial Implications {.smaller}
This clustering framework provides actionable guidance for retail managers:

- Fast-moving items: Prioritize to prevent stock-outs and protect revenue.

- Moderate-moving items: Maintain stable inventory and monitor for performance changes.

- Slow-moving items: Evaluate carefully to reduce waste, lower holding costs, and free shelf space.

Implementing K-Means clustering as part of inventory analytics helps retailers:

- Strengthen forecasting accuracy

- Improve inventory planning and allocation

- Reduce overstock and lost-sales risks

- Build a data-driven retail strategy grounded in objective performance patterns

Through this approach, retailers can better align supply with actual demand and enhance overall operational efficiency.

## Business Value {.smaller}

Using K-Means clustering on open government sales data:

Helps managers:

- Prioritize fast movers for availability and service level.

- Maintain balanced stocks for moderate movers.

- Control costs and space taken by slow movers.

Supports:

- Better demand forecasting

- More efficient inventory control

- Data-driven promotion and assortment decisions

Overall, clustering turns raw sales data into clear, actionable categories for retail decision-making.

## Limitations  {.smaller}

**Cluster Size Imbalance:**One drawback is that Cluster 2 is significantly larger than Clusters 1 and 3.  This reflects the fact that most retail items move slowly, but it also indicates that the clustering structure is severely uneven.  This could affect how well the smaller high-value clusters are represented.

**K-Means Assumptions:**K-Means assumes spherical, uniformly sized clusters and is sensitive to initial centroid location and outliers.  Given the skewed distribution of retail sales data, these assumptions may not always be valid.

**Limited Feature Set:**The analysis only considers sales-related factors (total sales, retail sales, and warehouse sales).  Important business parameters, such as profit margin, price elasticity, promotional activity, consumer demographics, and item lifetime, were not considered and could have an impact on cluster formation.

##  Future Work {.smaller}
**Test Alternative Clustering Algorithms:**Investigate models that can handle skewed and imbalanced data, such as DBSCAN, Gaussian Mixture Models, and hierarchical clustering, to see if a more balanced segmentation emerges.

**Add more business features:**Use margin, profitability, seasonality indications, promotions, or RFM-style customer KPIs to create clusters that reflect both sales volume and business value.

**Try Weighted or Hybrid models:**Consider weighted K-Means, MiniBatch K-Means, or hybrid models to reduce sensitivity to extreme values and improve resilience.

**Extend to customer-level clustering:**Using the same dataset, cluster consumer behavior (frequency, recency, spend, brand loyalty) to supplement product-level segmentation.

## References {.smaller}
